---
title: "Mega2GenABEL Vignette"
author: "Robert V. Baron and Daniel E. Weeks"
date: "`r format(Sys.time(), '%B %d, %Y, %R')`"
output:
  rmarkdown::html_vignette:
  pdf_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Mega2GenABEL Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

# Creating a GenABEL gwaa.data-class object using the mega2genabel R package

There are several data formats that Mega2 can read and transform
into a database that are not directly accepted by GenABEL. But any
Mega2 database that can be read into R, can be transformed to GenABEL
following this example using *seqsimr.db*.

GenABEL (http://www.genabel.org/) can process PLINK .tped/.tfam files with the function
*convert.snp.tped* to create a gwaa.data-class object.  Currently, we
use this mechanism to convert Mega2 data frames to a gwaa.data-class
object.  The argument to Mega2GenABEL is a file name prefix that will
be used to name the .tped/.fam files that are created.  These files
are read by GenABEL's *convert.snp.tped* (https://cran.r-project.org/package=GenABEL)
to produce a GenABEL "raw" file.  Then the "raw" file and a generated phenotype
file are processed by *load.gwaa.data* to yield a gwaa.data-class
object.

Type:

```{r mega2genabel}
library(mega2genabel)
#ENV = read.Mega2DB("seqsimr.db")

# or to use the embeded tutorial data
library(mega2rtutorial)
database_file = system.file("exdata", "seqsimr.db", package = "mega2rtutorial")
ENV = read.Mega2DB(database_file)

# This line converts the database to a gwaa.data-class object. The intermediate
# files begin with "SEQ"
seqsimgwaa = Mega2GenABEL("SEQ")
```

The files generated are:

```{r}
list.files(pattern = "SEQ.*")
```

You can use any of the GenABEL provided functions on the results, seqsimgwaa.

```{r str}
str(seqsimgwaa)
```

# GenABEL regression

This test is a bit more complicated.  We intend to verify that the object generated 
by the the previous process is the same as a GenABEL object we started with.
First, we load GenABEL and access its data:

```{r srdta}
library(GenABEL)
data(srdta)
```

Then we load the mega2genabel library and dump the GenABEL data as
PLINK .ped/.map/.phe files

```{r dmpPed}
library(mega2genabel)
dmpPed(gwaa_ = srdta, pfx = "srdta")
```

To produce the Mega2 database srdta.db, we switch to the Unix command line and run:


```{bash,eval=FALSE}
mega2 MEGA2.BATCH.srdta
```

**NOTE: To make this tutorial only dependent on R, the above code is not actually run.  And its
results, shown below, were captured from an environment where we had both R and the Mega2 executable
available.**

```
## ==========================================================
##                           MEGA2 4.9.2
##      Copyright (C) 1999-2017 Robert Baron, Justin R. Stickel, Charles P. Kollar,
##      Nandita Mukhopadhyay, Lee Almasy, Mark Schroeder, William P. Mulvihill,
##      Daniel E. Weeks, and University of Pittsburgh
## 
##      Last updated: Jun 13 2017, 09:36:42 , valid until June 15, 2018.
##      Compiled with gcc version 4.2.1 Compatible Apple LLVM 8.0.0 (clang-800.0.42.1)
## 
##      Mega2 comes with ABSOLUTELY NO WARRANTY.
##      See LICENSE.txt for terms of copying, modifying & redistributing Mega2.
## ==========================================================
## NOTE: For humans, chromosome 23 codes for X, 24 codes for Y and 25 codes for XY.
## 
## Run date:                  2017-7-24-10-05
## 
## Running Mega2 in batch mode from MEGA2.BATCH.srdta.
## Input filenames and missing value indicator read in from batch file.
## Dump Analysis option read in from batch file.
## WARNING: Locus selections not specified in batch file.
## WARNING: Going to Reorder menu.
## WARNING: Trait selections not specified in batch file.
## WARNING: Going to Trait selection menu.
## ==========================================================
## Keyword Input_Locus_File not in batch file, Locus file assumed to be unspecified.
## Keyword Input_Map_File not in batch file, Map file assumed to be unspecified.
## Keyword Input_Omit_File not in batch file, Omit file assumed to be unspecified.
## Keyword Input_Frequency_File not in batch file, Frequency file assumed to be unspecified.
## Keyword Input_Penetrance_File not in batch file, Penetrance file assumed to be unspecified.
## Keyword Input_Aux_File not in batch file, Aux file assumed to be unspecified.
## Keyword Input_Imputed_Info_File not in batch file, Imputed Info file assumed to be unspecified.
## ===========================================================
## Analysis Class: Dump.
## Quantitative  Input Missing Value  -9 
## Affection     Input Missing Value     "-9"
## Quantitative Output Missing Value      "*"
## Affection    Output Missing Value      "*"
## Input Format: PLINK PED format (ped)
## Pedigree and map files specified as PLINK format.
## omit, penetrance, and frequency files are always in Mega2 format.
## Input files will be read in as PLINK or Mega2 format files as appropriate.
## reading phenotype file srdta.phe ... (5 columns)
## Reading PLINK map file for names: srdta.map
## Reading map file srdta.map ... (4 columns)
## Input Map name: Map, type: average genetic map, units: kosambi Morgans
## Input Map name: BP, type: physical map
## Found 2 possible maps in the srdta.map file.
## Now checking each record in map file srdta.map ...
## Done reading map file: srdta.map
## 
## ===========================================================
## Total number of loci =  839
## 6 trait loci 
##       2 Affection status loci: 
##                 bt default
##       4 Quantitative loci: 
##                 age qt1 qt2 qt3
##       833 Marker loci 
## Number of loci found per chromosome (chromosome:number)
##    1:833
## ===========================================================
## WARNING: No frequency file provided.
## WARNING: Allele frequencies for these will be estimated from data.
## Trait 'bt' will be assigned the default penetrance: (0.0500 0.9000 0.9000)
## Trait 'default' will be assigned the default penetrance: (0.0500 0.9000 0.9000)
## Reading PLINK .ped file: srdta.ped (1672 columns).
## 833 (of 833) markers to be included from srdta.map
## Reading pedigree information from srdta.ped
## 2500 individuals read from srdta.ped
## 2500 individuals with nonmissing phenotypes
## 0 cases, 1244 controls, 1256 missing
## 1275 males, 1225 females, 0 of unspecified sex
## 0 founders, 2500 non-founders found
## ===========================================================
## Input pedigree data contains:
## Input pedigree file is in PLINK-fam format. 
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL     2500     2500    1275      1225       1978958        0    2082500
## Typed     2500     2500    1275      1225
## Untyped      0        0       0         0
## ===========================================================
## Pedigree exclusion option : Include all pedigrees whether typed or not.
## Count option: all alleles
## Count half-typed individuals' alleles : no 
## ===========================================================
## Recoding pedigree genotypes ... 
## ===========================================================
## Pedigree data summary after recoding:
## Input pedigree file is in PLINK-fam format. 
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL     2500     2500    1275      1225       1978958        0    2082500
## Typed     2500     2500    1275      1225
## Untyped      0        0       0         0
## ===========================================================
## Created linkage ped tree
## 
## ===== Messages of type "quant_stat": 
## ------------------------------------------------------------
## Per-pedigree quantitative phenotype summary:
## Pedigree     Mean      Std Dev     Minimum    Maximum  #Phenotypes
## ------------------------------------------------------------
## age
## ------------------------------------------------------------
## ------------------------------------------------------------
## p1           43.40000    0.00000   43.40000   43.40000        1
## p2           48.20000    0.00000   48.20000   48.20000        1
## p3           37.90000    0.00000   37.90000   37.90000        1
## p4           53.80000    0.00000   53.80000   53.80000        1
## p5           47.50000    0.00000   47.50000   47.50000        1
## p6           45.00000    0.00000   45.00000   45.00000        1
## p7           52.00000    0.00000   52.00000   52.00000        1
## p8           42.50000    0.00000   42.50000   42.50000        1
## p9           29.70000    0.00000   29.70000   29.70000        1
## ===== Too many "quant_stat" records, display is temporarily suspended ..
## qt1
## ------------------------------------------------------------
## ------------------------------------------------------------
## p1           -0.58000    0.00000   -0.58000   -0.58000        1
## p2            0.80000    0.00000    0.80000    0.80000        1
## p3           -0.52000    0.00000   -0.52000   -0.52000        1
## p4           -1.55000    0.00000   -1.55000   -1.55000        1
## p5            0.25000    0.00000    0.25000    0.25000        1
## p6            0.15000    0.00000    0.15000    0.15000        1
## p7           -0.56000    0.00000   -0.56000   -0.56000        1
## p8            0.00000    0.00000    0.00000    0.00000        0
## p9           -2.26000    0.00000   -2.26000   -2.26000        1
## ===== Too many "quant_stat" records, display is temporarily suspended ..
## qt2
## ------------------------------------------------------------
## ------------------------------------------------------------
## p1            4.46000    0.00000    4.46000    4.46000        1
## p2            6.32000    0.00000    6.32000    6.32000        1
## p3            3.26000    0.00000    3.26000    3.26000        1
## p4          888.00000    0.00000  888.00000  888.00000        1
## p5            5.70000    0.00000    5.70000    5.70000        1
## p6            4.65000    0.00000    4.65000    4.65000        1
## p7            4.64000    0.00000    4.64000    4.64000        1
## p8            5.77000    0.00000    5.77000    5.77000        1
## p9            0.71000    0.00000    0.71000    0.71000        1
## ===== Too many "quant_stat" records, display is temporarily suspended ..
## qt3
## ------------------------------------------------------------
## ------------------------------------------------------------
## p1            1.43000    0.00000    1.43000    1.43000        1
## p2            3.90000    0.00000    3.90000    3.90000        1
## p3            5.05000    0.00000    5.05000    5.05000        1
## p4            3.76000    0.00000    3.76000    3.76000        1
## p5            2.89000    0.00000    2.89000    2.89000        1
## p6            1.87000    0.00000    1.87000    1.87000        1
## p7            2.49000    0.00000    2.49000    2.49000        1
## p8            2.68000    0.00000    2.68000    2.68000        1
## p9            1.45000    0.00000    1.45000    1.45000        1
## ===== Too many "quant_stat" records, display is temporarily suspended ..
## ===== 2501 total records of type "quant_stat" are in MEGA2.LOG
## 
## 
## ===== Messages of type "quant_stat_sum": 
##                Quantitative trait phenotype statistics
## -------------------------------------------------------------------------------
## QTL            Missing  Minimum   Maximum   Total      Pedigrees    Total
##                                             pedigrees  phenotyped   phenotypes
## -------------------------------------------------------------------------------
## age                  0   24.100    71.600        2500        2500         2500
## qt1                  3   -4.600     3.200        2500        2497         2497
## qt2                  0    0.000   888.000        2500        2500         2500
## qt3                 11   -1.970     6.340        2500        2489         2489
## NOTE: The Missing QTL value on input has been assigned as '-9.00'.
## -------------------------------------------------------------------------
## QTL                 Mean   Std Dev  Skewness  Kurtosis
## -------------------------------------------------------------------------
## age               50.038     7.060     0.003    -0.063
## qt1               -0.298     1.001    -0.075     0.126
## qt2                6.122    30.601    28.734   825.077
## qt3                2.609     1.101     0.033    -0.090
## ===========================================================
## ===== 4 total records of type "quant_stat_sum" are in MEGA2.LOG
## 
## Done checking locus integrity.
## Checking pedigree integrity...
## Done checking pedigree integrity.
## ==========================================================
## ===========================================================
## Pedigree statistics after selecting chromosomes and marker loci:
## Input pedigree file is in post-makeped format.
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL     2500     2500    1275      1225       1978958        0    2082500
## Typed     2500     2500    1275      1225
## Untyped      0        0       0         0
## ===========================================================
## Database file "srdta.db" will be backed up.
## Moved existing srdta.db to srdta.db.old
## Dumping SQLite3 DB to file "srdta.db"
## ===========================================================
## See run summaries in directory 2017-7-24-10-05 
##    MEGA2.LOG, MEGA2.RECODE, MEGA2.ERR, MEGA2.KEYS
## The script 'mega2log2html.pl' exited normally.
## To view the HTML-formatted run summaries, open
## /Users/rbaron/mega2/bb/srcdir/R/mega2rtutorial/vignettes/2017-7-24-10-05/MEGA2run.html
## in a web browser.
## ===========================================================
```

## The MEGA2.BATCH.srdata file

The abbreviated MEGA2.BATCH.srdta file is below. (The initial comment section
is not shown and important lines are
shown in a bold typeface.)

> **Input_Database_Mode=1**  
**Input_Format_Type=4**  
**Input_Pedigree_File=srdta.ped**  
**Input_PLINK_Map_File=srdta.map**  
**Input_Phenotype_File=srdta.phe**  
Output_Path=.  
Input_Path=.  
PLINK_Args= --missing-phenotype -9 --trait default  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
Value_Marker_Compression=1  
**Analysis_Option=Dump**  
Value_Missing_Quant_On_Input=-9.000000  
Value_Missing_Affect_On_Input=-9  
Count_Genotypes=4  
Count_Halftyped=no  
Value_Genetic_Distance_Index=0  
Value_Genetic_Distance_SexTypeMap=0  
Value_Base_Pair_Position_Index=1  
Default_Reset_Invalid=no  
**DBfile_name=srdta.db**  
Default_Outfile_Names=yes  

Note: We have provided a copy of the srdta.db database in the mega2rtutorial.  So you can
proceed with the steps that follow, even if you did not create your own database.

So, we go back into R and type:

```{r srdtadb}
# ENV = read.Mega2DB("srdta.db")

# or to use the embeded tutorial data
library(mega2rtutorial)
database_file = system.file("exdata", "srdta.db", package = "mega2rtutorial")
ENV = read.Mega2DB(database_file)

mega = Mega2GenABEL("SRD", envir = ENV)
list.files(pattern = "SRD.*")
```

*mega* should be the same as the *srdta* gwaa.data-class object. You can
compare them however you prefer.  The function *Mega2GenABELtst()* can be used to
compare the phenotype data, genotype data, and object metadata, item by item.

In R, type:

```{r strs}
str(mega)
str(srdta)
```

to show the objects and then to compare them type:

```{r mega2genabeltst}
Mega2GenABELtst(mega_ = mega, gwaa_ = srdta)
```
