---
title: "Mega2r Vignette"
author: "Robert V. Baron and Daniel E. Weeks"
date: "`r format(Sys.time(), '%B %d, %Y, %R')`"
output:
  rmarkdown::html_vignette:
  pdf_document:
    highlight: pygments
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Mega2r Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

# Tutorial Data 

The files you will need for this tutorial are in the mega2rtutorial
package.  Most of the vignettes will access the data directly, but the
mega2rtutorial "vignettes" and the running of the "mega2" executable require
the files to be in the current directory.  This is done by running:

```{r,eval=FALSE}
library(mega2rtutorial)
dump_mega2rtutorial_data()
```

You should see the following files:  

```
--------------------------|-----------------------|---------------  
[1] "MEGA2.BATCH.seqsimr" |  "MEGA2.BATCH.srdta"  |  "MEGA2.BATCH.vcf"      
[4] "Mega2r.map"          |  "Mega2r.ped"         |  "mega2r.Rmd"   
[7] "mega2r.pdf"          |  "seqsimr.db"         |  "srdta.db"
```

The "clean" command will remove these files:

```{r, eval=FALSE}
clean_mega2rtutorial_data()
```

# Installation

Mega2 for R consist of the following 5 CRAN packages:  *mega2rtutorial*, *mega2r*, *mega2vcf*,
*mega2genabel* and *mega2pedgene*.
*mega2rtutorial* is needed to retrieve data and create a complete vignette/tutorial.
*mega2r* might be necessary and suffficent for your task.
*mega2vcf*, *mega2genabel* and *mega2pedgene* perform specific functions (see their package description);
these load the *mega2r* package when they are loaded.

Mega2r also requires that Bioconductor Annotations must be loaded.
The first line (below) loads the Bioconductor loader and the next
two lines load two annotation databases. One for gene transcript
locations and one to map gene names to the entrez gene IDs. (Note: You
may choose a different transcript database from Bioconductor or
construct one of your own.)

```
source("https://bioconductor.org/biocLite.R")
biocLite("TxDb.Hsapiens.UCSC.hg19.knownGene") 
biocLite("org.Hs.eg.db")
```

The above commands only need to be typed once.

# Example Data

We have chosen to use the SeqSIMLA2 program to generate a data set for
us.  We needed to sub-sample the data down to 1,380 people and 1,000
markers to make the size manageable. These data will be used to
illustrate Mega2 and Mega2R operations that follow.

Note: This simulator only produces markers on chromosome 1.

# Using Mega2R 

## Creating a Mega2 Database

We have provided files in the mega2rtutorial package, that contain
the data from the simulation. These files are in PLINK ped format.

* Mega2r.ped
* Mega2r.map

**This vignette will not invoke *mega2*, but use the seqsimr.db
database that is in the tutorial package.**

If you want, you can obtain the Mega2 program from
https://watson.hgen.pitt.edu/register/. Then you could invoke Mega2
on your data. To make matters simple, we will use a pre-constructed
Mega2 batch file to automate the processing by Mega2. To run Mega2 to
process and create the SQLite3 database 'seqsimr.db', issue this
command at the Unix prompt:

```
mega2 MEGA2.BATCH.seqsimr
```

**NOTE: To make this tutorial only dependent on R, the above code is not actually run.  And its
results, shown below, were captured from an environment where we had both R and Mega2 (the C++ program)
available.**


```
## ==========================================================
##                           MEGA2 4.9.2
##      Copyright (C) 1999-2017 Robert Baron, Justin R. Stickel, Charles P. Kollar,
##      Nandita Mukhopadhyay, Lee Almasy, Mark Schroeder, William P. Mulvihill,
##      Daniel E. Weeks, and University of Pittsburgh
## 
##      Last updated: Jun 13 2017, 09:36:42 , valid until June 15, 2018.
##      Compiled with gcc version 4.2.1 Compatible Apple LLVM 8.0.0 (clang-800.0.42.1)
## 
##      Mega2 comes with ABSOLUTELY NO WARRANTY.
##      See LICENSE.txt for terms of copying, modifying & redistributing Mega2.
## ==========================================================
## NOTE: For humans, chromosome 23 codes for X, 24 codes for Y and 25 codes for XY.
## 
## Run date:                  2017-7-24-10-03
## 
## Running Mega2 in batch mode from MEGA2.BATCH.seqsimr.
## Input filenames and missing value indicator read in from batch file.
## Dump Analysis option read in from batch file.
## WARNING: Locus selections not specified in batch file.
## WARNING: Going to Reorder menu.
## WARNING: Trait selections not specified in batch file.
## WARNING: Going to Trait selection menu.
## ==========================================================
## Keyword Input_Locus_File not in batch file, Locus file assumed to be unspecified.
## Keyword Input_Map_File not in batch file, Map file assumed to be unspecified.
## Keyword Input_Omit_File not in batch file, Omit file assumed to be unspecified.
## Keyword Input_Frequency_File not in batch file, Frequency file assumed to be unspecified.
## Keyword Input_Penetrance_File not in batch file, Penetrance file assumed to be unspecified.
## Keyword Input_Aux_File not in batch file, Aux file assumed to be unspecified.
## Keyword Input_Phenotype_File not in batch file, Phenotype file assumed to be unspecified.
## Keyword Input_Imputed_Info_File not in batch file, Imputed Info file assumed to be unspecified.
## ===========================================================
## Analysis Class: Dump.
## Quantitative  Input Missing Value  -9 
## Affection     Input Missing Value     "-9"
## Quantitative Output Missing Value      "*"
## Affection    Output Missing Value      "*"
## Input Format: PLINK PED format (ped)
## Pedigree and map files specified as PLINK format.
## omit, penetrance, and frequency files are always in Mega2 format.
## Input files will be read in as PLINK or Mega2 format files as appropriate.
## Reading PLINK map file for names: Mega2r.map
## Reading map file Mega2r.map ... (4 columns)
## Input Map name: Map, type: average genetic map, units: kosambi centiMorgans
## Input Map name: BP, type: physical map
## Found 2 possible maps in the Mega2r.map file.
## Now checking each record in map file Mega2r.map ...
## Done reading map file: Mega2r.map
## 
## ===========================================================
## Total number of loci =  1001
## 1 trait locus 
##       1 Affection status locus: 
##                 default
##       1000 Marker loci 
## Number of loci found per chromosome (chromosome:number)
##    1:1000
## ===========================================================
## WARNING: No frequency file provided.
## WARNING: Allele frequencies for these will be estimated from data.
## Trait 'default' will be assigned the default penetrance: (0.0500 0.9000 0.9000)
## Reading PLINK .ped file: Mega2r.ped (2006 columns).
## 1000 (of 1000) markers to be included from Mega2r.map
## Reading pedigree information from Mega2r.ped
## 1380 individuals read from Mega2r.ped
## 1380 individuals with nonmissing phenotypes
## 105 cases, 1275 controls, 0 missing
## 620 males, 760 females, 0 of unspecified sex
## 0 founders, 1380 non-founders found
## ===========================================================
## Input pedigree data contains:
## Input pedigree file is in PLINK-fam format. 
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL       20     1380     620       760       1380000        0    1380000
## Typed       20     1380     620       760
## Untyped      0        0       0         0
## ===========================================================
## Pedigree exclusion option : Include all pedigrees whether typed or not.
## Count option: all alleles
## Count half-typed individuals' alleles : no 
## ===========================================================
## Recoding pedigree genotypes ... 
## ===========================================================
## Pedigree data summary after recoding:
## Input pedigree file is in PLINK-fam format. 
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL       20     1380     620       760       1380000        0    1380000
## Typed       20     1380     620       760
## Untyped      0        0       0         0
## ===========================================================
## Created linkage ped tree
## Done checking locus integrity.
## Checking pedigree integrity...
## Done checking pedigree integrity.
## ==========================================================
## ===========================================================
## Pedigree statistics after selecting chromosomes and marker loci:
## Input pedigree file is in post-makeped format.
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL       20     1380     620       760       1380000        0    1380000
## Typed       20     1380     620       760
## Untyped      0        0       0         0
## ===========================================================
## Database file "seqsimr.db" will be backed up.
## Moved existing seqsimr.db to seqsimr.db.old
## Dumping SQLite3 DB to file "seqsimr.db"
## ===========================================================
## See run summaries in directory 2017-7-24-10-03 
##    MEGA2.LOG, MEGA2.RECODE, MEGA2.ERR, MEGA2.KEYS
## The script 'mega2log2html.pl' exited normally.
## To view the HTML-formatted run summaries, open
## /Users/rbaron/mega2/bb/srcdir/R/mega2rtutorial/vignettes/2017-7-24-10-03/MEGA2run.html
## in a web browser.
## ===========================================================
```

If you do not provide the argument, Mega2 will proceed to ask a series
of questions to collect the needed information to produce a
database. In addition, it will create a Mega2.BATCH file, we've
suggested you keep. You can look at the "Quick Start" documentation
https://watson.hgen.pitt.edu/docs/mega2_html/mega2.html Section 4. to
better understand the interactive process.

This MEGA2.BATCH.seqsimr file begins with a rather long comment
indicating the keyword values that may be set and their default
value. Toward the end of the file, we set the inputs as **Mega2r.ped**
and **Mega2r.map**, indicate the input is PLINK ped format with
parameters, and indicate that Mega2 should produce a database called
**seqsimr.db**, etc. (These particular items are in bold face text
below.)

> **Input_Database_Mode=1**  
**Input_Format_Type=4**  
**Input_Pedigree_File=Mega2r.ped**  
**Input_PLINK_Map_File=Mega2r.map**  
Output_Path=.  
Input_Path=.  
**PLINK_Args= --cM --missing-phenotype -9 --trait default**  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
Value_Marker_Compression=1  
**Analysis_Option=Dump**  
Value_Missing_Quant_On_Input=-9.000000  
Value_Missing_Affect_On_Input=-9  
Count_Genotypes=4  
Count_Halftyped=no  
Value_Genetic_Distance_Index=0  
Value_Genetic_Distance_SexTypeMap=0   
Value_Base_Pair_Position_Index=1  
Default_Reset_Invalid=no  
**DBfile_name=seqsimr.db**  
Default_Outfile_Names=yes  

You will eventually have to run Mega2 to convert your data.


## Reading and Examining a Mega2 Database in R

### Reading a Mega2 Database into R

After you have the database, start up the R program. The first R commands you should type are:

```{r,eval=FALSE}
library(mega2r)
ENV = read.Mega2DB("seqsimr.db", verbose = TRUE)
```

The first argument should be the path to the database. Providing the
optional argument, verbose, causes the read function to summarize the
tables created, their fields and their sizes. Finally, an "R
environment" that contains the database tables is returned. (If you
are unfamiliar with environments, you can think of them as data
frames. If fact, ENV$locus_table will access the locus_table variable
from ENV as though fetching an "observation" from a data frame. The
difference is when you change a data frame passed to a function, the
change does not affect the original data frame. Only the function's
local value is changed; ALL changes are forgotten when the function
exits. When you change the data in an environment passed to a function,
the change is permanent and seen inside and outside the function.)

```{r mega2r}
library(mega2r)
# ENV = read.Mega2DB("seqsimr.db", verbose = TRUE)
# or to use the embeded tutorial data
library(mega2rtutorial)
database_file = system.file("exdata", "seqsimr.db", package = "mega2rtutorial")
ENV = read.Mega2DB(database_file)
```

We need to make two observations that apply to all Mega2R functions:

### Verbose Flag

When verbose is set in the initial read.Mega2DB, the
value will be remembered.  It may be used by any subsequent function.
If verbose is TRUE, Mega2R functions will possibly print out more
diagnostic information.

### Use of an Environment

All Mega2R functions that do not return an
environment, need to have an environment supplied as an argument.
As stated earlier the environment is used to store the data frames that
contain the database data.
There are two ways to pass the environment.  If you assigned the result of
read.Mega2DB to the variable *seqsimr*, then you could supply the value
*seqsimr* to any Mega2R function as the named argument **envir**, viz.

```
showMega2ENV(envir = seqsimr)
```

The second choice is a bit of a "hack" but it is very
convenient. Every Mega2R function (that does not return an
environment) has a named argument defined as

```
envir = ENV
```

as in

```
showMega2ENV = function(envir = ENV) { ... }
```

The code above, sets global variable **ENV** to the local 
variable, envir.  And the argument evaluation mechanism of R will supports
this usage; even though **ENV** is not known until runtime.

## Reading Examining a Mega2 Database in R

The ls function will show you all the variables in an
environment. (You probably used it without arguments to show you the
variables in the .GlobalEnv.) Type:

```{r}
ls(ENV)
```


### A more informative overview of the data base can be seen by:

```{r}
showMega2ENV() 
```

### Use standard R commands to examine the created data frames

```{r}
str(ENV$locus_table)
```

# Iterating through the Transcripts

## Running pedgene on Transcripts

There are two ways to compute a function on all the genotypes (or markers) in a transcript.
(These examples are explained more thoroughly in the mega2pedgene vignette.)

### applyFnToRanges

Mega2R has an internal default list of the chromosome and base pair
ranges for a number of gene transcripts.  These transcripts come from
the UCSC Genome Browser reference assembly GRCH37.  The list was
further modified to eliminate multiple records of the same gene with
the exact same transcript start and transcript end.  This constitutes
about 29,000 records.  You may load you own range set instead of the default.
You create a data frame that contains at least a name "observation", a chromosome
"observation", a start position "observation" and an end position "observation".
You create an integer vector that contains the column numbers of the chromosome
"observation", the start position "observation" and the end position "observation".
These two become the arguments to the *setRanges* function, viz.
```
        setRanges(Transcripts, Offsets)
```
The function:

> applyFnToRanges(DOpedgene, envir = ENV)  

goes through each transcript/range entry and finds the markers that fall within the bounds.
It then builds a genotype matrix for the samples and each marker.

The first argument of the *applyFnToRanges* function is a call back function.  The function is
called with three arguments: the genotype matrix, the markers, and the selected transcript/range
entry.  The call back is invoked repeatedly for each transcript range that has any markers present.
If it is necessary to store information between successive invocations the environment (*envir*) 
can be used.
In mega2pedgene, DOpedgene converts the genotype
nucleotides to the count of the minor allele and then runs pedgene for
several different marker weights. 

*applyFnToRanges* can also be explicitly provided ranges.
```
applyFnToRanges(DOpedgene, ENV$refRanges, ENV$refIndices, envir = ENV)
applyFnToRanges(DOpedgene, ENV$refRanges[ENV$refRanges$chromosome == 1,], ENV$refIndices, envir = ENV)
```

### applyFnToGenes

Alternatively, you may try searching by gene.  Here the default
transcript database is TxDb.Hsapiens.UCSC.hg19.knownGene from
Bioconductor.  Of course you can change it.  Type:

```
setAnnotations("txdb", "genedb")
```

where "txdb" is a string that is the name of transcript database that was
fetched from Bioconductor, and similarly "genedb" is the name of a Bioconductor
database that maps a gene id from the transcript to an entrez gene id.
The *setAnnotations* command must be issued whenever R is started and the "mega2r" library has been loaded.

The function:

> applyFnToGenes(DOpedgene, genes_arg = c('DISP1', 'KIF26B', 'AK5', 'ST7L'), envir = ENV)

will look up all the transcripts for each gene in the Bioconductor database.  It will then behave
like *applyFnToRanges*, find all the markers in each range and then invoke the call back function with
the required 3 arguments.

*applyFnToGenes* has several other arguments that allow complete chromosomes, or (multiple) ranges of 
base pairs along chromosomes to be specified.

# Iterating through the Genotypes

## Cooked Genotypes

The meat of the above two functions is the calculation of the matrix of samples cross markers
containing the nucleotides (strings).  The
genotype information is most likely in a compressed 2 bit representation.  A Rcpp function does the
heavy lifting to convert the compressed genotype data to nucleotides.  The function

> getgenotypes(markers, envir = ENV)

returns the matrix for the specified markers.  It can take one additional argument to indicate a string to 
separate the two alleles of each nucleotide.

## Raw Genotypes

The function

> getgenotypesraw(markers, envir = ENV)

is similar to the *getgenotypes* function except that the matrix it returns contains integers.  The integers are 
encoded so that high 16 bits of the integer are the index for allele1 and the low 16 bits are the index for allele2.


# Next steps

If you are left with questions as to how you can use Mega2R for your own research, you can:

1. Read our paper "The Mega2R suite of R packages: tools for accessing
and processing common genetic data formats in R", which extends this
document by providing implementation details.

2. Read the source for the packages if you understand R; the code is not particularly subtle.

3. Write to the Mega2 discussion group: https://groups.google.com/forum/#!forum/mega2-users
