---
title: "Mega2RSKAT Tutorial"
author: "Robert V. Baron and Daniel E. Weeks"
date: "`r format(Sys.time(), '%B %d, %Y, %R')`"
output:
  rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Mega2RSKAT Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

  pdf_document:
    toc: true
    number_sections: true
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```
# 1 Introduction

## 1.1 What is Mega2

The Mega2 executable transforms genetic data available in several formats:
LINKAGE format, PLINK format, IMPUTE format and VCF format, into an 'SQLite' database.
For an association or linkage analysis project, Mega2 takes the database file and
transforms it into various other file formats required by the analysis programs,
thus greatly facilitating a variety of different analyses. Mega2 is described in,
https://watson.hgen.pitt.edu/docs/mega2_html/mega2.html.
(The source for Mega2 as well as Mega2R can be found at
https://bitbucket.org/dweeks/mega2 BitBucket site.)

## 1.2 What is Mega2R

Mega2R loads and manipulates data frames containing genotype, phenotype, marker,
and family information from an input 'SQLite' database.  In addition, it allows one
to select ranges defined as standard transcripts or as a data frame with start position,
end position and chromosome and process genotypes/markers of each range.  Arbitrary functions
of the genotype can be evaluated for each range.

## 1.3 What is SKAT

SKAT is a CRAN project (https://CRAN.R-project.org/package=SKAT) that
provides Snp-set (sequence) Kernel Association Tests.  It contains
"Functions for kernel-regression-based association tests including
Burden test, SKAT and SKAT-O. These methods aggregate individual SNP
score statistics in a SNP set and efficiently compute SNP-set level
p-values."  The Vignettes file (https://cran.r-project.org/web/packages/SKAT/vignettes/SKAT.pdf)
shows how to apply various test to a collection of markers.

## 1.4 What is Mega2RSKAT

Mega2RSKAT uses the capabilities of Mega2R to apply the SKAT functionality to all the 
transcripts.  This is analogous to how Mega2Rpedgene extend the CRAN pedgene
package.

# 2 Example Genome-wide Association Data

We used the SeqSIMLA2 program to generate an example data set to use in this vignette.
[SeqSIMLA2: simulating correlated quantitative traits accounting for shared
environmental effects in user-specified pedigree structure, Chung RH1,
Tsai WY, Hsieh CH, Hung KY, Hsiung CA, Hauser ER., Genet Epidemiol.
2015 Jan;39(1):20-4. doi: 10.1002/gepi.21850. Epub 2014 Sep 22.]
We needed to sub-sample the data down to 1,380 people and 1,000
markers to make the size manageable. These data will be used to
illustrate Mega2 and Mega2R operations that follow.

Note: The simulator produces markers on only chromosome 1.

# 3 Tutorial Data

The files you will need for this tutorial are provided in this package.
Further, our use of the "mega2" executable expects the Mega2.BATCH.<name> files to be in the
working directory and the latter files expect their data files to be in the
working directory.  This is done by running:

```{r mega2rtutDump, message=FALSE}
library(Mega2R)
dump_mega2rtutorial_data()
```

You should see the following files:  

```
--------------------------|-----------------------|---------------  
[1] "MEGA2.BATCH.seqsimr" |  "MEGA2.BATCH.srdta"  |  "MEGA2.BATCH.vcf"      
[4] "Mega2r.map"          |  "Mega2r.ped"         |  "mega2r.Rmd"   
[7] "mega2r.pdf"          |  "seqsimr.db"         |  "srdta.db"   
```

When you are done with these exercise, the "clean" command will remove these files:

```{r, mega2rtutClean,eval=FALSE}
clean_mega2rtutorial_data()
```


# 4 Using Mega2R to carry out automated gene-based association tests using 'SKAT'

The 'SKAT' package implements methods for carrying out gene-based association tests; 
it is available on CRAN as (https://CRAN.R-project.org/package=SKAT).
It was written by Seunggeun Lee and Michael Wu: 
[Lee, S., Emond, M.J., Bamshad, M.J., Barnes, K.C., Rieder, M.J., Nickerson, D.A., NHLBI GO
Exome Sequencing Project-ESP Lung Project Team, Christiani, D.C., Wurfel, M.M. and Lin, X.
(2012) Optimal unified approach for rare variant association testing with application to small sample
case-control whole-exome sequencing studies. American Journal of Human Genetics, 91, 224-237.]

Mega2R provides functions that permit one to run the 'SKAT' package to carry
out gene-based association tests using a kernel regression framework while
looping over selected marker subsets.


## 4.1 Loading a Mega2 database

Rather than read the Mega2 'SQLite' database with the customary `read.Mega2DB` function,
we use the custom `init_SKAT` function to read the Mega2 database and initialize for processing.
This function calls the utility function, `dbmega2_import`, to read the data base.
Then it creates, edits, and rewrites the family data, storing it in the data frame, `fam`. 
(`fam` contains data merged from the `pedigree_table`, `person_table` and
`phenotype_table`.)
`init_SKAT` purges persons from the `fam` data frame with unknown case/control
status. `setfam` that sets `fam` in the environment insures that
similar filtering is done to the phenotype_table and the genotype_table.
Next, `init_SKAT` decodes the phenotype_table into a simple data frame, `phe`.
You will definitely want to examine `ENV`$`phe` to choose which phenotype to use in the 
study as the case/control.
In addition, `init_SKAT` calculates
some values that will be used repeatedly and stores them in the environment that is returned.
Finally, `init_SKAT` stores the argument `allMarkers` which tells later processing to ignore
markers that show no variation (if FALSE).

```{r init_SKAT,message=FALSE}
ENV = init_SKAT("seqsimr.db", filename = "SKAT.txt", verbose = T, allMarkers = F)
```

## 4.2 Gene transcripts reminder

Genes are defined according to a database containing the boundaries of the gene transcripts
or defined by an internal table, `refRanges` with the boundaries.
The gene transcripts require two Bioconductor Annotations databases to be installed.
The first line (below) loads the Bioconductor loader and the next
two lines install two annotation databases. One annotation database provides the gene transcript
locations and the other maps gene names to entrez gene IDs. (Note: You
may choose a different transcript database from Bioconductor or
construct one of your own.) Please type in R:

```
source("https://bioconductor.org/biocLite.R")

biocLite("TxDb.Hsapiens.UCSC.hg19.knownGene") 
biocLite("org.Hs.eg.db")
```
The above step is run once.

Mega2R presumes that the databases, "TxDb.Hsapiens.UCSC.hg19.knownGene" and "org.Hs.eg.db" are 
selected.  Otherwise you must make your choices known to Mega2 via the command:
```
setAnnotations(txdb, entrezGene)
```
where `txdb` is the name of Bioconductor transcription database, and `entrezGene` is the name of Bioconductor mapping of gene name or gene alias to entrez gene id.

## 4.3 Gene Ranges reminder

Mega2R also has an internal default list of the chromosome and base pair
ranges for many gene transcripts.  These transcripts come from
the UCSC Genome Browser reference assembly GRCH37.  The list was
further modified to eliminate multiple records from the same gene with
the exact same transcript start and transcript end.  The list contains
about 29,000 records.  You may load your own range set instead of the default.
You create a data frame that contains at least a name "observation", a chromosome
"observation", a start position "observation" and an end position "observation".
And you create an integer vector that contains the column numbers of the chromosome
"observation", the start position "observation" and the end position "observation".
These two become the arguments to the `setRanges` function, viz.
```
        setRanges(Ranges, Columns)
```

## 4.4 Running SKAT on transcripts

You should be familiar with the SKAT functions in CRAN SKAT before you read this section.
The function, `Mega2SKAT`, is Mega2's basic interface to SKAT, both to `SKAT_Null_Model` and
`SKAT`.  It's signature is:

Mega2SKAT = function (f, ty, gs = 1:100, skat = SKAT::SKAT, envir = ENV, ...) { }

The `gs` argument indicates how many range elements should be processed and the `envir`
argument specifies the environment that contains all the Mega2R data frames.

Most of the time, before you call `SKAT`, you need to call `SKAT_Null_Model` with a formula
and an indicator for the type of the phenotype.  `Mega2SKAT` will take its first two arguments,
a formula and a type (string) and call `SKAT_Null_Model` with this information, viz.
          SKAT_Null_Model(f, out_type = ty)
If the formula, `f`, is NULL, Mega2SKAT will not call `SKAT_Null_Model` and you must do the 
equivalent before calling `Mega2SKAT`.  There are several reasons a custom call to the build
the model could be necessary.  You might want to use `SKAT_Null_Model` but provided additional 
arguments viz. data, Adjustment, n.Resampling, type.Resampling.  Alternatively, you might
need to use a different model viz. SKAT_NULL_emmaX, SKAT_Null_Model_ChrX.

The `skat` argument specifies the name of the SKAT package function to use; this is 
commonly SKAT::SKAT, but could be SKAT::SKATBinary, SKAT::SKAT_CommonRare, etc.
Additional, arguments needed for these "skat" functions are provided to the `Mega2SKAT`
function and will be passed to the eventual call.  All the "skat" functions are called with
a genotype matrix, an object representing the Null Model and possibly some additional arguments.
If the genotype matrix is empty, Mega2SKAT will issue a warning.  If some of markers have no 
variation, Mega2SKAT will ignore them if the variable allMarkers is FALSE.  If it is TRUE,
Mega2SKAT will allow them, but the SKAT will more than likely issue a warning.

The Mega2R loop engine, iterates through each range and determines the set of markers contained.
Mega2SKAT (`DOSKAT`) calculates a genotype matrix from the markers of a range and calls the 
specified "skat" function with the required arguments.  `DOSKAT` mainly converts the genotype
information of the form 1/1, 1/2, and 2/2 to 0, 1, and 2 with the Major allele being 0.

A typical invocation of `Mega2SKAT` could be:
Mega2SKAT(ENV$phe[, 3] - 1 ~ 1, "D", kernel = "linear.weighted", weights.beta=c(0.5,0.5))
The function `Mega2SKAT` examines the first 100 transcripts and prints the results.
Note: `verbose` needs to be TRUE, for these diagnostics to be printed.
```
CEP104 snp24037 520 646 214  
CEP104 snp24039 940 386 54  
CEP104 snp24041 1377 3 0  
CEP104 snp24048 860 460 60  
CEP104 snp24494 789 501 90  
CEP104 snp24499 891 442 47  
CEP104 snp24506 891 442 47  
CEP104 snp24507 789 501 90
```
These data are saved in the data frame, `SKAT_results`, in the environment.
Note: The results are always appended to the data frame.  You should truncate it
when necessary.

```{r, FnToRanges,eval=FALSE}
# we will skip this line for the Rmd document production because it takes too long
applyFnToRanges(DOSKAT, ENV$refRanges, ENV$refIndices, envir = ENV)
```

## 4.5 Running SKAT on transcripts

```{r,eval=FALSE}
applyFnToGenes(DOSKAT, genes_arg = c('DISP1', 'KIF26B', 'AK5', 'ST7L'), envir = ENV)
```

But let us run this function for a few genes:

```{r,FnToGenes,eval=TRUE}
applyFnToGenes(DOSKAT, genes_arg = c('DISP1', 'AK5'), envir = ENV)
```

Note: The default behavior of `DOSKAT` is to append any new results to the end of the
`ENV$SKAT_results` data frame.

NOTE: You will observe that for several genes there are a few
transcripts with the same or overlapping start values and end values.

# 5 Acknowlegements

The Mega2 C++ program, the Mega2R R package, and this package,
Mega2RSKAT are open source and are freely available.  SKAT is also
open source found in CRAN.  All the R packages can be found in
CRAN. This work was supported by NIH grant R01 GM076667 (PI: Weeks).
