---
title: "Mega2R Tutorial"
author: "Robert V. Baron and Daniel E. Weeks"
date: "`r format(Sys.time(), '%B %d, %Y, %R')`"
output:
  rmarkdown::html_vignette:
  pdf_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Mega2r Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

# Tutorial directory and files 

The files you will need for this tutorial are in the mega2rtutorial
package.  The "mega2" executable requires the Mega2.BATCH.name files to be in the
working directory; to copy these files to the working directory run:

```{r mega2rtutDump}
library(mega2rtutorial)
dump_mega2rtutorial_data()
```

You should see the following files:  

```
........................  | ....................  | ...............  
--------------------------|-----------------------|---------------  
[1] "MEGA2.BATCH.seqsimr" |  "MEGA2.BATCH.srdta"  |  "MEGA2.BATCH.vcf"      
[4] "Mega2r.map"          |  "Mega2r.ped"         |  "mega2r.Rmd"   
[7] "mega2r.pdf"          |  "seqsimr.db"         |  "srdta.db"   
```

When you are done with these exercise, the "clean" command will remove these files:

```{r, mega2rtutClean,eval=FALSE}
clean_mega2rtutorial_data()
```
Note: The vignettes for the other packages take an alternate approach to using
the data directly from the mega2rtutorial package without moving them to the
working directory.

# Installation

**Note: The code in this section should only be executed once, to load
the packages into R and are NOT run for every R session.**

**Also, we will assume that you have started an R session at which to type the commands
in the tutorial.**

To run these
exercises, you should install the CRAN packages `mega2r`, `mega2vcf`,
`mega2genabel` and `mega2pedgene`. Once they are in CRAN, the
preferred way to install will be to use the install.packages()
command. For now, we prefer to use devtools in R; as in

```
devtools::install("mega2rtutorial")
devtools::install("mega2r")
devtools::install("mega2pedgene")
devtools::install("mega2vcf")
devtools::install("mega2genabel")
```

Certainly, issuing these commands at the UNIX prompt:

```
R CMD install mega2rtutorial
R CMD install mega2r
R CMD install mega2pedgene  
R CMD install mega2vcf  
R CMD install mega2genabel  
```

will work, too. They may generate warnings about non-package related
items in the tree. Some of these packages may fetch other dependent CRAN
packages.

In addition, we require two Bioconductor Annotations to be loaded.
The first line (below) loads the Bioconductor loader and the next
two lines install two annotation databases. One for gene transcript
locations and one to map gene names to entrez gene IDs. (Note: You
may choose a different transcript database from Bioconductor or
construct one of your own.) Please type in R:

```
source("https://bioconductor.org/biocLite.R")
biocLite("TxDb.Hsapiens.UCSC.hg19.knownGene") 
biocLite("org.Hs.eg.db")
```


# Example Data

We have chosen to use the SeqSIMLA2 program to generate a data set for
us.  We needed to sub-sample the data down to 1,380 people and 1,000
markers to make the size manageable. These data will be used to
illustrate Mega2 and Mega2R operations that follow.

Note: This simulator only produces markers on chromosome 1.

# Using Mega2R packages

## Creating a Mega2 database

We have provided files in the mega2rtutorial package, that contain
the data from the simulation. These files are in PLINK ped format data
with 1,380 samples of 1,000 markers each:

* Mega2r.ped
* Mega2r.map

**If you do not wish to install Mega2 right now, you can use the seqsimr.db
database that is in the tutorial directory.  Of course, the tutorial steps
that invoke Mega2 will fail.**

You will need to obtain the Mega2 program from
https://watson.hgen.pitt.edu/register/. Then, you will invoke Mega2
on your data. To make matters simple, we will use a pre-constructed
Mega2 batch file to automate the processing by Mega2. To run Mega2 to
process and create the SQLite3 database 'seqsimr.db', we issue this
command at the Unix prompt:

```{bash, eval=FALSE}
mega2 MEGA2.BATCH.seqsimr
```

**NOTE: To make this tutorial only dependent on R, the above code is not actually run.  And its
results, shown below, were captured from an environment where we had both R and Mega2 executable
available.  All the examples of mega2 shown in these exercises have been similarly "fudged".**

```
## ==========================================================
##                           MEGA2 4.9.2
##      Copyright (C) 1999-2017 Robert Baron, Justin R. Stickel, Charles P. Kollar,
##      Nandita Mukhopadhyay, Lee Almasy, Mark Schroeder, William P. Mulvihill,
##      Daniel E. Weeks, and University of Pittsburgh
## 
##      Last updated: Jun 13 2017, 09:36:42 , valid until June 15, 2018.
##      Compiled with gcc version 4.2.1 Compatible Apple LLVM 8.0.0 (clang-800.0.42.1)
## 
##      Mega2 comes with ABSOLUTELY NO WARRANTY.
##      See LICENSE.txt for terms of copying, modifying & redistributing Mega2.
## ==========================================================
## NOTE: For humans, chromosome 23 codes for X, 24 codes for Y and 25 codes for XY.
## 
## Run date:                  2017-7-24-10-03
## 
## Running Mega2 in batch mode from MEGA2.BATCH.seqsimr.
## Input filenames and missing value indicator read in from batch file.
## Dump Analysis option read in from batch file.
## WARNING: Locus selections not specified in batch file.
## WARNING: Going to Reorder menu.
## WARNING: Trait selections not specified in batch file.
## WARNING: Going to Trait selection menu.
## ==========================================================
## Keyword Input_Locus_File not in batch file, Locus file assumed to be unspecified.
## Keyword Input_Map_File not in batch file, Map file assumed to be unspecified.
## Keyword Input_Omit_File not in batch file, Omit file assumed to be unspecified.
## Keyword Input_Frequency_File not in batch file, Frequency file assumed to be unspecified.
## Keyword Input_Penetrance_File not in batch file, Penetrance file assumed to be unspecified.
## Keyword Input_Aux_File not in batch file, Aux file assumed to be unspecified.
## Keyword Input_Phenotype_File not in batch file, Phenotype file assumed to be unspecified.
## Keyword Input_Imputed_Info_File not in batch file, Imputed Info file assumed to be unspecified.
## ===========================================================
## Analysis Class: Dump.
## Quantitative  Input Missing Value  -9 
## Affection     Input Missing Value     "-9"
## Quantitative Output Missing Value      "*"
## Affection    Output Missing Value      "*"
## Input Format: PLINK PED format (ped)
## Pedigree and map files specified as PLINK format.
## omit, penetrance, and frequency files are always in Mega2 format.
## Input files will be read in as PLINK or Mega2 format files as appropriate.
## Reading PLINK map file for names: Mega2r.map
## Reading map file Mega2r.map ... (4 columns)
## Input Map name: Map, type: average genetic map, units: kosambi centiMorgans
## Input Map name: BP, type: physical map
## Found 2 possible maps in the Mega2r.map file.
## Now checking each record in map file Mega2r.map ...
## Done reading map file: Mega2r.map
## 
## ===========================================================
## Total number of loci =  1001
## 1 trait locus 
##       1 Affection status locus: 
##                 default
##       1000 Marker loci 
## Number of loci found per chromosome (chromosome:number)
##    1:1000
## ===========================================================
## WARNING: No frequency file provided.
## WARNING: Allele frequencies for these will be estimated from data.
## Trait 'default' will be assigned the default penetrance: (0.0500 0.9000 0.9000)
## Reading PLINK .ped file: Mega2r.ped (2006 columns).
## 1000 (of 1000) markers to be included from Mega2r.map
## Reading pedigree information from Mega2r.ped
## 1380 individuals read from Mega2r.ped
## 1380 individuals with nonmissing phenotypes
## 105 cases, 1275 controls, 0 missing
## 620 males, 760 females, 0 of unspecified sex
## 0 founders, 1380 non-founders found
## ===========================================================
## Input pedigree data contains:
## Input pedigree file is in PLINK-fam format. 
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL       20     1380     620       760       1380000        0    1380000
## Typed       20     1380     620       760
## Untyped      0        0       0         0
## ===========================================================
## Pedigree exclusion option : Include all pedigrees whether typed or not.
## Count option: all alleles
## Count half-typed individuals' alleles : no 
## ===========================================================
## Recoding pedigree genotypes ... 
## ===========================================================
## Pedigree data summary after recoding:
## Input pedigree file is in PLINK-fam format. 
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL       20     1380     620       760       1380000        0    1380000
## Typed       20     1380     620       760
## Untyped      0        0       0         0
## ===========================================================
## Created linkage ped tree
## Done checking locus integrity.
## Checking pedigree integrity...
## Done checking pedigree integrity.
## ==========================================================
## ===========================================================
## Pedigree statistics after selecting chromosomes and marker loci:
## Input pedigree file is in post-makeped format.
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL       20     1380     620       760       1380000        0    1380000
## Typed       20     1380     620       760
## Untyped      0        0       0         0
## ===========================================================
## Database file "seqsimr.db" will be backed up.
## Moved existing seqsimr.db to seqsimr.db.old
## Dumping SQLite3 DB to file "seqsimr.db"
## ===========================================================
## See run summaries in directory 2017-7-24-10-03 
##    MEGA2.LOG, MEGA2.RECODE, MEGA2.ERR, MEGA2.KEYS
## The script 'mega2log2html.pl' exited normally.
## To view the HTML-formatted run summaries, open
## /Users/rbaron/mega2/bb/srcdir/R/mega2rtutorial/vignettes/2017-7-24-10-03/MEGA2run.html
## in a web browser.
## ===========================================================
```

If you do not provide an argument, Mega2 will proceed to ask a series
of questions to collect the information needed to produce a
database. In addition, it will create a Mega2.BATCH file, we
suggested you use. You can look at the "Quick Start" section of the Mega2 documentation
https://watson.hgen.pitt.edu/docs/mega2_html/mega2.html to
better understand the interactive process.

The MEGA2.BATCH.seqsimr file begins with a rather long comment
indicating the keyword values that may be set and their default
value. Toward the end of the file, we see the inputs set to **Mega2r.ped**
and **Mega2r.map**, indicate the input is PLINK ped format with
parameters, and indicate that Mega2 should produce a database called
**seqsimr.db**, etc. (These particular items are in bold face text
below.)

> **Input_Database_Mode=1**  
**Input_Format_Type=4**  
**Input_Pedigree_File=Mega2r.ped**  
**Input_PLINK_Map_File=Mega2r.map**  
Output_Path=.  
Input_Path=.  
**PLINK_Args= --cM --missing-phenotype -9 --trait default**  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
Value_Marker_Compression=1  
**Analysis_Option=Dump**  
Value_Missing_Quant_On_Input=-9.000000  
Value_Missing_Affect_On_Input=-9  
Count_Genotypes=4  
Count_Halftyped=no  
Value_Genetic_Distance_Index=0  
Value_Genetic_Distance_SexTypeMap=0   
Value_Base_Pair_Position_Index=1  
Default_Reset_Invalid=no  
**DBfile_name=seqsimr.db**  
Default_Outfile_Names=yes  

If you wish to use any of the Mega2R suite of R packages, you will eventually have to run Mega2 to convert your data into the required SQLite database.


## Reading a Mega2 database into R

After you have the database, start up the R program. The first R commands you should type are:

```{r, seqsimrdb,eval=FALSE}
library(mega2r)
ENV = read.Mega2DB("seqsimr.db", verbose = TRUE)
```

The first argument should be the path to the database. Providing the
optional argument, `verbose`, causes the read function to summarize the
tables created, their fields and their sizes. Finally, an "R
environment", `ENV`, that contains the database tables is returned. (If you
are unfamiliar with environments, you can think of them as data
frames. `ENV$locus_table` will access the `locus_table` variable
from `ENV` similar to fetching an "observation" from a data frame. The
difference is when you change a data frame passed to a function, the
change does not affect the original data frame. Only the function's
local value is changed; ALL changes are forgotten when the function
exits. If you change the data in an environment passed to a function,
the change is permanent.)

```{r seqsimrdb2}
library(mega2r)
ENV = read.Mega2DB("seqsimr.db", verbose = TRUE)
```

We need to make two observations that apply to all Mega2R functions:
The first is that when verbose is set in the initial read.Mega2DB, the
value will be remembered.  It may be used by any subsequent function.
If verbose is TRUE, Mega2R functions can print diagnostic information.
The second is that all Mega2R functions that do not return an
environment, need to have an environment supplied as an argument.
There are two ways to accomplish this.  If you assigned the result of
`read.Mega2DB` to the variable `seqsimr`, then you could supply the value
seqsimr to any Mega2R function as the named argument `envir`:

```
showMega2ENV(envir = seqsimr)
```

The second choice is a bit of a "hack" but it is very
convenient. Every Mega2R function (that does not return an
environment) has a named `envir` argument defined to take on the default value `ENV`: 

```
envir = ENV
```

as in

```
showMega2ENV = function(envir = ENV) { ... }
```

The code above, sets global variable `ENV` to the local 
variable, `envir`.  When `envir` is evaluated in the function, it 
will look up the value of `ENV` in the global environment.
Hopefully, `ENV` was set to the return value of `read.Mega2DB`

### Back to more examples.

The `ls` function will show you all the variables in an
environment. (You probably have used it without arguments to show you the
variables in the .GlobalEnv.) Type:

```{r, ls}
ls(ENV)
```


A more informative overview of the data base can be had with:

```{r, showMega2ENV}
showMega2ENV() 
```

### Use standard R commands to examine the created data frames

```{r, str}
str(ENV$locus_table)
```

## Running pedgene using the mega2pedgene R package

### Loading a Mega2 database

Rather than use
the `read.Mega2DB` function described previously and used by all the other
packages, this package uses `init_pedgene` to read the Mega2 database.
This function calls a utility function
shared with `read.Mega2DB`. Then it creates, edits, and rewrites the family data -- 
data frame, `fam` .  (`fam` merges data from the `pedigree_table`, `person_table` and
`phenotype_table`.)
`init_pedgene` purges persons with unknown case/control
status which is necessary for the pedgene calculation.
(When `fam` is filtered, similar adjustments are made to the phenotype_table and the
genotype_table.)
Finally, `init_pedgene` calculates
some values that will be used repeatedly and stores them in the environment that is returned.

```{r init_pedgene}
library(mega2pedgene)
ENV = init_pedgene("seqsimr.db", verbose = TRUE)
```

### Running pedgene on transcripts

Mega2R has an internal default list of the chromosome and base pair
ranges for a number of gene transcripts.  These transcripts come from
the UCSC Genome Browser reference assembly GRCH37.  The list was
further modified to eliminate multiple records of the same gene with
the exact same transcript start and transcript end.  These data contain
about 29,000 records.   You may load your own range set instead of the default.
You create a data frame that contains at least a name "observation", a chromosome
"observation", a start position "observation" and an end position "observation".
Then you create an integer vector that contains the column numbers of the chromosome
"observation", the start position "observation" and the end position "observation".
These two objects are then arguments to the `setRanges` function, viz.
```
        setRanges(Transcripts, Columns)
```

By default, the function run_pedgene examines the first 100
transcripts and prints the results. For this database, the first 100 transcripts
identifies one usable transcript with several markers (It is usable 
because it is on chromosome 1 and it has a range that overlaps with some of the
markers in our data.)
To make this exercise run 
faster, we noticed that this transcript appeared at transcript 54; so we will
restrict pedgene to a small range of transcripts around 54, viz. 50 through
60: 

```{r run_pedgene,collapse=TRUE}
run_pedgene(gs=50:60)
```

You will see many reports
of "No markers in range", but occasionally you will see a listing of a
gene name, markers, and count of 1/1, 1/2 and 2/2 genotypes, viz.

```
CEP104 snp24037 520 646 214  
CEP104 snp24039 940 386 54  
CEP104 snp24041 1377 3 0  
CEP104 snp24048 860 460 60  
CEP104 snp24494 789 501 90  
CEP104 snp24499 891 442 47  
CEP104 snp24506 891 442 47  
CEP104 snp24507 789 501 90
```

The genotypes for these markers, along with data frame of these markers, the range
used, and the environment are fed
to the call back function `DOpedgene`. `DOpedgene` converts the genotype
nucleotides to 0, 1, and 2 to represent the genotype
and then runs pedgene for
several different marker weights. The results are automatically stored
in a data frame with "observations": prefix of chromosome, gene,
number of markers and base pair range followed by Pedgene data: kernel
and burden, value and p-values, four values for each of three
weightings of the markers.
These data are saved in the data frame, `pedgene_results`, in the environment.
They are also printed when `verbose` is TRUE, viz.

```
   chr   gene nvariants   start     end sKernel_BT pKernel_BT sBurden_BT  
1 chr1 CEP104         8 3728644 3773797   31.96998  0.6297541 -0.6496837  
  pBurden_BT sKernel_MB pKernel_MB sBurden_MB pBurden_MB sKernel_UW pKernel_UW  
1  0.5158965   1184.995  0.5138866  -1.209563  0.2264468   222.8737  0.4111136  
  sBurden_UW pBurden_UW  
1  -1.169488   0.242207
```

You should run pedgene on all the transcript entries, but it takes a rather long time.
So type:

```{r, FnToRanges,eval=FALSE}
# we will skip this line for the Rmd document production because it takes too long
applyFnToRanges(DOpedgene, ENV$refRanges, ENV$refIndices, envir = ENV)
```

If you run the above test, you will see that gene DISP1 and KIF26B
have at least one p-value less than 0.01 and AK5 and STL7 at least one less than 0.03.

### Running pedgene on selected genes

You may try searching for transcripts of specific genes.  Here, the default
transcript database is `TxDb.Hsapiens.UCSC.hg19.knownGene` from
Bioconductor.  Of course you can change it.  Type:

```
setAnnotations("txdb", "genedb")
```

where "txdb" is a string that is the name of transcript database that was
fetched from Bioconductor, and similarly "genedb" is the name of a Bioconductor
database that maps a gene id from the input to an entrez gene id.
You need to install any new the database with biocLite, as shown earlier.

We leave the command below as an exercise.  It runs a bit slowly.  It needs to find
all  the transcripts for each gene; then it needs to find all the markers between each pair
of transcript start/end ranges; and finally it needs to do a small computation on the genotypes
from each set of markers.
```{r,eval=FALSE,collapse=TRUE}
applyFnToGenes(DOpedgene, genes_arg = c('DISP1', 'KIF26B', 'AK5', 'ST7L'), envir = ENV)
```

But let us run this function for a few genes:

```{r,FnToGenes,eval=TRUE,collapse=TRUE}
applyFnToGenes(DOpedgene, genes_arg = c('DISP1', 'AK5'), envir = ENV)
```

Note that the default behavior of `DOpedgene` is to append any new results to the end of the
`ENV$pedgene_results` data frame.

NOTE: You will observe that for several genes there are a few
transcripts with the same or overlapping start values and end values.

## Writing a VCF File using the mega2vcf R package

First, load the mega2vcf library and then create the directory "vcfr", so we can keep all the generated
data in one place.

```{r mega2vcflib}
library(mega2vcf)
if (! dir.exists("vcfr")) dir.create("vcfr")
```

The Mega2VCF function takes an argument which is the prefix used on all
generated files. We will make that a path that includes the
directory *vcfr*.  The Mega2VCF function assumes the Mega2 data frames
are contained in the global variable ENV. If this is not the case, supply a
named argument "envir" set to the environment you wish to use.

Note, that Mega2VCF will create the .vcf file as well as a
.fam file (we do linkage), a .phe file (phenotypes) and a few
others. These files are listed below.

Creating the VCF file and related files is accomplished by typing:

```{r mega2vcf}
Mega2VCF("vcfr/vcf.01", ENV$markers[ENV$markers$chromosome==1 ,])
```

Note: The line above places the data for only chromosome 1 in the files 
(Recall that our simulated data is only on chromosome 1).
You can filter `ENV$markers` as you wish.
If the second argument is not present, all the markers are written out.

The generated files are listed below:
```{r, lsvcf}
list.files("vcfr")
```

You can read the R code to see how these files are created from the data frames.


## Creating a GenABEL gwaa.data-class object using the mega2genabel R package


There are several data formats that Mega2 can read and transform
into a database that are not directly accepted by GenABEL. But any
Mega2 database that can be read into R, can be transformed to GenABEL
following this example using *seqsimr.db*.

GenABEL (http://www.genabel.org/) can process PLINK .tped/.tfam files with the function
`convert.snp.tped` to create a gwaa.data-class object.  Currently, we
use this mechanism to convert Mega2 data frames to a gwaa.data-class
object.  The argument to Mega2GenABEL is a file name prefix that will
be used to name the .tped/.fam files that are created.  These files
are read by GenABEL's `convert.snp.tped` (https://cran.r-project.org/package=GenABEL)
to produce a GenABEL "raw" file.  Then the "raw" file and a generated phenotype
file are processed by `load.gwaa.data` to yield a gwaa.data-class
object.

If you haven't already, first install the `GenABEL` R package from CRAN.

Then type:

```{r seqsimgwaa}
library(mega2genabel)
ENV = read.Mega2DB("seqsimr.db")

# This line converts the database to a gwaa.data-class object. The intermediate
# files begin with "SEQ"
seqsimgwaa = Mega2GenABEL("SEQ")
```

The files generated are:

```
list.files(pattern = "SEQ.*")
```

You can use any of the GenABEL provided functions on the results, seqsimgwaa.

```{r strseqsimgwaa}
str(seqsimgwaa)
```

# Regression

To verify that the R packages described above have worked correctly,
we can compare their output files to those created by Mega2 itself, to
verify that they are identical, as expected. This type of testing is
known as regression testing.

## VCF regression

If you have a compiled copy of Mega2, you can run this code using
the MEGA2.BATCH.vcf file and database from the mega2rtutorial package.
Cut and paste the shell code and you should see similar results to what
is presented here.
Otherwise, the text below is an illustration of what should happen.
These lines use the C++ Mega2 program to populate the vcf directory with
the same set of files that the R program created in the vcfr
directory.  

At the Unix command prompt, type:

```{bash, eval=FALSE}
mkdir vcf
mega2 MEGA2.BATCH.vcf
```

```
## ==========================================================
##                           MEGA2 4.9.2
##      Copyright (C) 1999-2017 Robert Baron, Justin R. Stickel, Charles P. Kollar,
##      Nandita Mukhopadhyay, Lee Almasy, Mark Schroeder, William P. Mulvihill,
##      Daniel E. Weeks, and University of Pittsburgh
## 
##      Last updated: Jun 13 2017, 09:36:42 , valid until June 15, 2018.
##      Compiled with gcc version 4.2.1 Compatible Apple LLVM 8.0.0 (clang-800.0.42.1)
## 
##      Mega2 comes with ABSOLUTELY NO WARRANTY.
##      See LICENSE.txt for terms of copying, modifying & redistributing Mega2.
## ==========================================================
## NOTE: For humans, chromosome 23 codes for X, 24 codes for Y and 25 codes for XY.
## 
## Run date:                  2017-7-24-10-05
## 
## Running Mega2 in batch mode from MEGA2.BATCH.vcf.
## Analysis option read in from batch file.
## Chromosome(s) and markers read in from batch file.
## Trait selection(s) read in from batch file.
## ==========================================================
## Analysis Class: VCF.
## Quantitative Output Missing Value     "-9"
## Affection    Output Missing Value     "-9"
## Reading SQLite3 DB from file "seqsimr.db"
## ===========================================================
## The path to this SQLite3 database is seqsimr.db.
## This database was created using Mega2 version 4.9.2.
## This database was created using   SQLite3 3.9.2 on 2017-7-24-10-03.
## This database was processed using SQLite3 3.9.2 on 2017-7-24-10-05.
## This database was created from PLINK PED format data using the following files:
## 	       Pedigree file   Mega2r.ped
## 	      PLINK Map file   Mega2r.map
## This database contains:
## 	1380 persons (20 pedigrees)
## 	1000 markers
## 	1 trait
## 	genetic distance(map name/type) "Map"/kosambi, Sex map type AVERAGED_MAP
## 	base pair distance(map name) "BP"
## 
## ==========================================================
## 1 trait locus 
##       1 Affection status locus: 
##                 default
## ===========================================================
## Selected map Map.
## Selected chromosome 1
## Output will combine markers and the following selected traits:
##                 default [MARKERS]
## After selecting traits and covariates
## 1 trait locus 
##       1 Affection status locus: 
##                 default
## ===========================================================
## Pedigree statistics after selecting chromosomes and marker loci:
## Input pedigree file is in post-makeped format.
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL       20     1380     620       760       1380000        0    1380000
## Typed       20     1380     620       760
## Untyped      0        0       0         0
## ===========================================================
## Mega2 created the following file(s) for VCF Format:
## VCF file created using allele ordering setting: Original_Order
##         VCF format file:        vcf/vcf.01.vcf
##         VCF pedigree file:      vcf/vcf.01.fam
##         VCF phenotype file:     vcf/vcf.01.phe
##         VCF map file:           vcf/vcf.01.map
##         VCF freq file:          vcf/vcf.01.freq
##         VCF pen file:           vcf/vcf.01.pen
## 
## SQLite3 database "seqsimr.db" was processed to generate this output.
## Output is in vcf
## ===========================================================
## See run summaries in directory 2017-7-24-10-05 
##    MEGA2.LOG, MEGA2.ERR, MEGA2.KEYS
## The script 'mega2log2html.pl' exited normally.
## To view the HTML-formatted run summaries, open
## /Users/rbaron/mega2/bb/srcdir/R/mega2rtutorial/vignettes/2017-7-24-10-05/MEGA2run.html
## in a web browser.
## ===========================================================
## If you use Mega2 as part of a published work, please reference 
##  Baron RV, Kollar C, Mukhopadhyay N, Weeks DE
##  Mega2: validated data-reformatting for linkage and association analyses
##  Source Code for Biology and Medicine.2014, 9:26
##  DOI: 10.1186/s13029-014-0026-y
## as well as the version used, which is currently Version 4.9.2
## ===========================================================
```


The abbreviated MEGA2.BATCH.vcf file is below (notice that there are
no INPUT FILES, just a database file):

> Input_Database_Mode=2  
Align_Strand_Input=no  
**Output_Path=vcf**  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
**Analysis_Option=VCF**  
Value_Missing_Quant_On_Output=-9  
Value_Missing_Affect_On_Output=-9  
**DBfile_name=seqsimr.db**  
Chromosome_Single=1  
Traits_Combine=1 2 e  
**file_name_stem=vcf**  
human_genome_build=B37  
VCF_output_file_type=1  
VCF_Allele_Order=Original_Order  
Default_Outfile_Names=yes  

You should compare the two directories: vcf and vcfr. Please add the
-w flag to the diff command. This causes the comparison to ignore
differences in white space. You may may notice that expressions that
contain dates that are different between the two directories.  Type:

```{bash, diffvcf,eval=FALSE}
diff -wrs vcf vcfr
```

```
## Files vcf/vcf.01.fam and vcfr/vcf.01.fam are identical
## Files vcf/vcf.01.freq and vcfr/vcf.01.freq are identical
## Files vcf/vcf.01.map and vcfr/vcf.01.map are identical
## Files vcf/vcf.01.pen and vcfr/vcf.01.pen are identical
## Files vcf/vcf.01.phe and vcfr/vcf.01.phe are identical
## Files vcf/vcf.01.vcf and vcfr/vcf.01.vcf are identical
```

## GenABEL regression

This test is a bit more complicated. First, we load GenABEL and access its data:

```{r srdta}
library(GenABEL)
data(srdta)
```

Then we load the `mega2genabel` library and dump the GenABEL data as
PLINK .ped/.map/.phe files

```{r dmpPed}
library(mega2genabel)
dmpPed(gwaa_ = srdta, pfx = "srdta")
```

To produce the Mega2 database `srdta.db`, we switch to the Unix command line and run:

```{bash, eval=FALSE}
mega2 MEGA2.BATCH.srdta
```

**NOTE: To make this tutorial only dependent on R, the above code is not actually run.  And its
results, shown below, were captured from an environment where we had both R and the Mega2 executable
available.**

```
## ==========================================================
##                           MEGA2 4.9.2
##      Copyright (C) 1999-2017 Robert Baron, Justin R. Stickel, Charles P. Kollar,
##      Nandita Mukhopadhyay, Lee Almasy, Mark Schroeder, William P. Mulvihill,
##      Daniel E. Weeks, and University of Pittsburgh
## 
##      Last updated: Jun 13 2017, 09:36:42 , valid until June 15, 2018.
##      Compiled with gcc version 4.2.1 Compatible Apple LLVM 8.0.0 (clang-800.0.42.1)
## 
##      Mega2 comes with ABSOLUTELY NO WARRANTY.
##      See LICENSE.txt for terms of copying, modifying & redistributing Mega2.
## ==========================================================
## NOTE: For humans, chromosome 23 codes for X, 24 codes for Y and 25 codes for XY.
## 
## Run date:                  2017-7-24-10-05
## 
## Running Mega2 in batch mode from MEGA2.BATCH.srdta.
## Input filenames and missing value indicator read in from batch file.
## Dump Analysis option read in from batch file.
## WARNING: Locus selections not specified in batch file.
## WARNING: Going to Reorder menu.
## WARNING: Trait selections not specified in batch file.
## WARNING: Going to Trait selection menu.
## ==========================================================
## Keyword Input_Locus_File not in batch file, Locus file assumed to be unspecified.
## Keyword Input_Map_File not in batch file, Map file assumed to be unspecified.
## Keyword Input_Omit_File not in batch file, Omit file assumed to be unspecified.
## Keyword Input_Frequency_File not in batch file, Frequency file assumed to be unspecified.
## Keyword Input_Penetrance_File not in batch file, Penetrance file assumed to be unspecified.
## Keyword Input_Aux_File not in batch file, Aux file assumed to be unspecified.
## Keyword Input_Imputed_Info_File not in batch file, Imputed Info file assumed to be unspecified.
## ===========================================================
## Analysis Class: Dump.
## Quantitative  Input Missing Value  -9 
## Affection     Input Missing Value     "-9"
## Quantitative Output Missing Value      "*"
## Affection    Output Missing Value      "*"
## Input Format: PLINK PED format (ped)
## Pedigree and map files specified as PLINK format.
## omit, penetrance, and frequency files are always in Mega2 format.
## Input files will be read in as PLINK or Mega2 format files as appropriate.
## reading phenotype file srdta.phe ... (5 columns)
## Reading PLINK map file for names: srdta.map
## Reading map file srdta.map ... (4 columns)
## Input Map name: Map, type: average genetic map, units: kosambi Morgans
## Input Map name: BP, type: physical map
## Found 2 possible maps in the srdta.map file.
## Now checking each record in map file srdta.map ...
## Done reading map file: srdta.map
## 
## ===========================================================
## Total number of loci =  839
## 6 trait loci 
##       2 Affection status loci: 
##                 bt default
##       4 Quantitative loci: 
##                 age qt1 qt2 qt3
##       833 Marker loci 
## Number of loci found per chromosome (chromosome:number)
##    1:833
## ===========================================================
## WARNING: No frequency file provided.
## WARNING: Allele frequencies for these will be estimated from data.
## Trait 'bt' will be assigned the default penetrance: (0.0500 0.9000 0.9000)
## Trait 'default' will be assigned the default penetrance: (0.0500 0.9000 0.9000)
## Reading PLINK .ped file: srdta.ped (1672 columns).
## 833 (of 833) markers to be included from srdta.map
## Reading pedigree information from srdta.ped
## 2500 individuals read from srdta.ped
## 2500 individuals with nonmissing phenotypes
## 0 cases, 1244 controls, 1256 missing
## 1275 males, 1225 females, 0 of unspecified sex
## 0 founders, 2500 non-founders found
## ===========================================================
## Input pedigree data contains:
## Input pedigree file is in PLINK-fam format. 
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL     2500     2500    1275      1225       1978958        0    2082500
## Typed     2500     2500    1275      1225
## Untyped      0        0       0         0
## ===========================================================
## Pedigree exclusion option : Include all pedigrees whether typed or not.
## Count option: all alleles
## Count half-typed individuals' alleles : no 
## ===========================================================
## Recoding pedigree genotypes ... 
## ===========================================================
## Pedigree data summary after recoding:
## Input pedigree file is in PLINK-fam format. 
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL     2500     2500    1275      1225       1978958        0    2082500
## Typed     2500     2500    1275      1225
## Untyped      0        0       0         0
## ===========================================================
## Created linkage ped tree
## 
## ===== Messages of type "quant_stat": 
## ------------------------------------------------------------
## Per-pedigree quantitative phenotype summary:
## Pedigree     Mean      Std Dev     Minimum    Maximum  #Phenotypes
## ------------------------------------------------------------
## age
## ------------------------------------------------------------
## ------------------------------------------------------------
## p1           43.40000    0.00000   43.40000   43.40000        1
## p2           48.20000    0.00000   48.20000   48.20000        1
## p3           37.90000    0.00000   37.90000   37.90000        1
## p4           53.80000    0.00000   53.80000   53.80000        1
## p5           47.50000    0.00000   47.50000   47.50000        1
## p6           45.00000    0.00000   45.00000   45.00000        1
## p7           52.00000    0.00000   52.00000   52.00000        1
## p8           42.50000    0.00000   42.50000   42.50000        1
## p9           29.70000    0.00000   29.70000   29.70000        1
## ===== Too many "quant_stat" records, display is temporarily suspended ..
## qt1
## ------------------------------------------------------------
## ------------------------------------------------------------
## p1           -0.58000    0.00000   -0.58000   -0.58000        1
## p2            0.80000    0.00000    0.80000    0.80000        1
## p3           -0.52000    0.00000   -0.52000   -0.52000        1
## p4           -1.55000    0.00000   -1.55000   -1.55000        1
## p5            0.25000    0.00000    0.25000    0.25000        1
## p6            0.15000    0.00000    0.15000    0.15000        1
## p7           -0.56000    0.00000   -0.56000   -0.56000        1
## p8            0.00000    0.00000    0.00000    0.00000        0
## p9           -2.26000    0.00000   -2.26000   -2.26000        1
## ===== Too many "quant_stat" records, display is temporarily suspended ..
## qt2
## ------------------------------------------------------------
## ------------------------------------------------------------
## p1            4.46000    0.00000    4.46000    4.46000        1
## p2            6.32000    0.00000    6.32000    6.32000        1
## p3            3.26000    0.00000    3.26000    3.26000        1
## p4          888.00000    0.00000  888.00000  888.00000        1
## p5            5.70000    0.00000    5.70000    5.70000        1
## p6            4.65000    0.00000    4.65000    4.65000        1
## p7            4.64000    0.00000    4.64000    4.64000        1
## p8            5.77000    0.00000    5.77000    5.77000        1
## p9            0.71000    0.00000    0.71000    0.71000        1
## ===== Too many "quant_stat" records, display is temporarily suspended ..
## qt3
## ------------------------------------------------------------
## ------------------------------------------------------------
## p1            1.43000    0.00000    1.43000    1.43000        1
## p2            3.90000    0.00000    3.90000    3.90000        1
## p3            5.05000    0.00000    5.05000    5.05000        1
## p4            3.76000    0.00000    3.76000    3.76000        1
## p5            2.89000    0.00000    2.89000    2.89000        1
## p6            1.87000    0.00000    1.87000    1.87000        1
## p7            2.49000    0.00000    2.49000    2.49000        1
## p8            2.68000    0.00000    2.68000    2.68000        1
## p9            1.45000    0.00000    1.45000    1.45000        1
## ===== Too many "quant_stat" records, display is temporarily suspended ..
## ===== 2501 total records of type "quant_stat" are in MEGA2.LOG
## 
## 
## ===== Messages of type "quant_stat_sum": 
##                Quantitative trait phenotype statistics
## -------------------------------------------------------------------------------
## QTL            Missing  Minimum   Maximum   Total      Pedigrees    Total
##                                             pedigrees  phenotyped   phenotypes
## -------------------------------------------------------------------------------
## age                  0   24.100    71.600        2500        2500         2500
## qt1                  3   -4.600     3.200        2500        2497         2497
## qt2                  0    0.000   888.000        2500        2500         2500
## qt3                 11   -1.970     6.340        2500        2489         2489
## NOTE: The Missing QTL value on input has been assigned as '-9.00'.
## -------------------------------------------------------------------------
## QTL                 Mean   Std Dev  Skewness  Kurtosis
## -------------------------------------------------------------------------
## age               50.038     7.060     0.003    -0.063
## qt1               -0.298     1.001    -0.075     0.126
## qt2                6.122    30.601    28.734   825.077
## qt3                2.609     1.101     0.033    -0.090
## ===========================================================
## ===== 4 total records of type "quant_stat_sum" are in MEGA2.LOG
## 
## Done checking locus integrity.
## Checking pedigree integrity...
## Done checking pedigree integrity.
## ==========================================================
## ===========================================================
## Pedigree statistics after selecting chromosomes and marker loci:
## Input pedigree file is in post-makeped format.
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL     2500     2500    1275      1225       1978958        0    2082500
## Typed     2500     2500    1275      1225
## Untyped      0        0       0         0
## ===========================================================
## Database file "srdta.db" will be backed up.
## Moved existing srdta.db to srdta.db.old
## Dumping SQLite3 DB to file "srdta.db"
## ===========================================================
## See run summaries in directory 2017-7-24-10-05 
##    MEGA2.LOG, MEGA2.RECODE, MEGA2.ERR, MEGA2.KEYS
## The script 'mega2log2html.pl' exited normally.
## To view the HTML-formatted run summaries, open
## /Users/rbaron/mega2/bb/srcdir/R/mega2rtutorial/vignettes/2017-7-24-10-05/MEGA2run.html
## in a web browser.
## ===========================================================
```

### The MEGA2.BATCH.srdata file

The abbreviated MEGA2.BATCH.srdta file is below. (The initial comment section
is not shown and important lines are shown in a bold typeface.)

> **Input_Database_Mode=1**  
**Input_Format_Type=4**  
**Input_Pedigree_File=srdta.ped**  
**Input_PLINK_Map_File=srdta.map**  
**Input_Phenotype_File=srdta.phe**  
Output_Path=.  
Input_Path=.  
PLINK_Args= --missing-phenotype -9 --trait default  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
Value_Marker_Compression=1  
**Analysis_Option=Dump**  
Value_Missing_Quant_On_Input=-9.000000  
Value_Missing_Affect_On_Input=-9  
Count_Genotypes=4  
Count_Halftyped=no  
Value_Genetic_Distance_Index=0  
Value_Genetic_Distance_SexTypeMap=0  
Value_Base_Pair_Position_Index=1  
Default_Reset_Invalid=no  
**DBfile_name=srdta.db**  
Default_Outfile_Names=yes  


Note: We have provided a copy of the srdta.db database in the mega2rtutorial.  So you can
proceed with the steps that follow, even if you did not create your own database.

So, we go back into R and type:

```{r, srdta.db,eval=TRUE}
ENV = read.Mega2DB("srdta.db")
mega = Mega2GenABEL("SRD")
list.files(pattern = "SRD.*")
```

`mega` should be the same as the `srdta` gwaa.data-class object. You can
compare them however you prefer.  But, `Mega2GenABELtst()` can be used to
compare the phenotype data, genotype data, and object metadata, item by item.

In R, type:

```{r strs}
str(mega)
str(srdta)
```

to show the objects and then to compare them type:

```{r, mega2genabeltst,eval=TRUE}
options(max.print = 30)
Mega2GenABELtst(mega_ = mega, gwaa_ = srdta)
```

# Next steps

If you are left with questions as to how you can use Mega2R for your own research, you can:

1. Read our paper "The Mega2R suite of R packages: tools for accessing
and processing common genetic data formats in R", which extends this
document by providing implementation details.

2. Read the source for the packages if you understand R; the code is not particularly subtle.

3. Write to the Mega2 discussion group: https://groups.google.com/forum/#!forum/mega2-users

```{r, fin,eval=TRUE,echo=FALSE}
mega2rtutorial::clean_mega2rtutorial_data()
```
