---
title: "Example usage of Mega2R"
author: "Robert V. Baron and Daniel E. Weeks"
date: "`r format(Sys.time(), '%B %d, %Y, %R')`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
require(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

# Tutorial directory and files 

The files you will need for this tutorial are in the mega2rtutorial
directory.  You should **cd** there to do these exercises.:

> You should see the following files:  
........................  | ....................  | ...............  
--------------------------|-----------------------|---------------  
[1] "MEGA2.BATCH.seqsimr" |  "MEGA2.BATCH.srdta"  |  "MEGA2.BATCH.vcf"      
[4] "Mega2r.map"          |  "Mega2r.ped"         |  "mega2r.Rmd"   
[7] "mega2r.pdf"          |  "seqsimr.db"         |    


# Installation

**Note: The code in this section should only be executed once, to load
the packages into R and are NOT run for every R session.**

**Also, we will assume that you have started an R session at which to type the commands
in the tutorial.**

To run these
exercises, you should install the CRAN packages *mega2r*, *mega2vcf*,
*mega2genabel* and *mega2pedgene*. Once they are in CRAN, the
preferred way to install will be to use the install.packages()
command. For now, we prefer to use devtools in R; as in

```{r,cache=TRUE,eval=FALSE}
devtools::install("../mega2r")
devtools::install("../mega2pedgene")
devtools::install("../mega2vcf")
devtools::install("../mega2genabel")
```

Certainly, 

> R CMD install mega2r  
R CMD install mega2pedgene  
R CMD install mega2vcf  
R CMD install mega2genabel  

will work, too. It may generate warnings about non-package related
items in the tree. All these packages may fetch other dependent CRAN
packages. We also require Bioconductor Annotations that must be loaded
manually. The first line loads the Bioconductor loader and the next
two lines load two annotation databases. One for gene transcript
locations and one to map gene names to the entrez gene IDs. (Note: You
may choose a different transcript database from Bioconductor or
construct one of your own.) Please type in R:

```{r,cache=TRUE,eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("TxDb.Hsapiens.UCSC.hg19.knownGene") 
biocLite("org.Hs.eg.db")
```


# Example Data

We have chosen to use the SeqSIMLA2 program to generate a data set for
us.  We needed to sub-sample the data down to 1,380 people and 1,000
markers to make the size manageable. These data will be used to
illustrate Mega2 and Mega2R operations that follow.

Note: This simulator only produces markers on chromosome 1.

# Using Mega2R packages

## Creating a Mega2 database

We have provided files in the mega2rtutorial directory, that contain
the data from the simulation. These files are in PLINK ped format data
for 1,380 samples with 1,000 markers:

* seqsimr.ped
* seqsimr.map

**If you do not wish to install Mega2 right now, you can use the seqsimr.db
database that is in the tutorial directory.  Of course, the tutorial steps
that invoke Mega2 will fail, mainly the regression steps.**

You will need to obtain the Mega2 program from
https://watson.hgen.pitt.edu/register/. First, you will invoke Mega2
on your data. To make matters simple, we will use a pre-constructed
Mega2 batch file to automate the processing by Mega2. To run Mega2 to
process and create the SQLite3 database 'seqsimr.db', we issue this
command at the Unix prompt:

```{bash,results="hide"}
mega2 MEGA2.BATCH.seqsimr
```

If you do not provide the argument, Mega2 will proceed to ask a series
of questions to collect the needed information to produce a
database. In addition, it will create a Mega2.BATCH file, we've
suggested you use. You can look at the "Quick Start" documentation
https://watson.hgen.pitt.edu/docs/mega2_html/mega2.html Section 4. to
better understand the interactive process.

This MEGA2.BATCH.seqsimr file begins with a rather long comment
indicating the keyword values that may be set and their default
value. Toward the end of the file, we set the inputs as **Mega2r.ped**
and **Mega2r.map**, indicate the input is PLINK ped format with
parameters, and indicate that Mega2 should produce a database called
**seqsimr.db**, etc. (These particular items are in bold face text
below.)

> **Input_Database_Mode=1**  
**Input_Format_Type=4**  
**Input_Pedigree_File=Mega2r.ped**  
**Input_PLINK_Map_File=Mega2r.map**  
Output_Path=.  
Input_Path=.  
**PLINK_Args= --cM --missing-phenotype -9 --trait default**  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
Value_Marker_Compression=1  
**Analysis_Option=Dump**  
Value_Missing_Quant_On_Input=-9.000000  
Value_Missing_Affect_On_Input=-9  
Count_Genotypes=4  
Count_Halftyped=no  
Value_Genetic_Distance_Index=0  
Value_Genetic_Distance_SexTypeMap=0   
Value_Base_Pair_Position_Index=1  
Default_Reset_Invalid=no  
**DBfile_name=seqsimr.db**  
Default_Outfile_Names=yes  

You will eventually have to run Mega2 to convert your data.


## Reading a Mega2 database into R

After you have the database, start up the R program. The first R commands you should type are:

```{r,eval=FALSE}
library(mega2r)
ENV = read.Mega2DB("seqsimr.db", verbose = TRUE)
```

The first argument should be the path to the database. Providing the
optional argument, verbose, causes the read function to summarize the
tables created, their fields and their sizes. Finally, an "R
environment" that contains the database tables is returned. (If you
are unfamiliar with environments, you can think of them as data
frames. If fact, ENV$locus_table will access the locus_table variable
from ENV as though fetching an "observation" from a data frame. The
difference is when you change a data frame passed to a function, the
change does not affect the original data frame. Only the function's
local value is changed; ALL changes are forgotten when the function
exits. If you change the data in an environment passed to a function,
the change is permanent.)

```{r}
library(mega2r)
ENV = read.Mega2DB("seqsimr.db", verbose = TRUE)
```

We need to make two observations that apply to all Mega2R functions:
The first is that when verbose is set in the initial read.Mega2DB, the
value will be remembered.  It may be used by any subsequent function.
If verbose is TRUE, Mega2R functions will possibly print out far more
diagnostic information than you might need, especially on subsequent
runs.  The second is that all Mega2R functions that do not return an
environment, need to have an environment supplied as an argument.
There are two ways to accomplish this.  If you assigned the result of
read.Mega2DB to the variable seqsimr, then you could supply the value
seqsimr to any Mega2R function as the named argument **envir**:

> showMega2ENV(envir = seqsimr)

The second choice is a bit of a "hack" but it is very
convenient. Every Mega2R function (that does not return an
environment) has a named argument defined as

> envir = ENV

as in

> showMega2ENV = function(envir = ENV) { ... }

The code above, sets global variable **ENV** to the local 
variable, envir.  And the argument evaluation mechanism of R will supports
this usage; even though **ENV** is not known until runtime.

### Back to more examples.

The ls function will show you all the variables in an
environment. (You probably used it without arguments to show you the
variables in the .GlobalEnv.) Type:

```{r}
ls(ENV)
```


### A more informative overview of the data base can be seen by:

```{r}
showMega2ENV() 
```

### Use standard R commands to examine the created data frames

```{r}
str(ENV$locus_table)
```

## Running pedgene using the mega2pedgene R package

mega2pedgene is the exception that proves the rule. Rather than use
the read.Mega2DB function described above and as used by all the other
packages, we use init_pedgene. This function calls a utility function
created for read.Mega2DB. Then it creates, edits, and rewrites the
**fam** data frame to purge persons with unknown case/control
status. (**fam** merges data from the pedigree_table, person_table and
phenotype_table as you might expect.)  Finally, it calculates and
stores some values that will be used repeatedly. viz:

```{r}
library(mega2pedgene)
ENV = init_pedgene("seqsimr.db", verbose = T)
```

Mega2R has an internal default list of the chromosome and base pair
ranges for a number of gene transcripts.  These transcripts come from
the UCSC Genome Browser reference assembly GRCH37.  The list was
further modified to eliminate multiple records of the same gene with
the exact same transcript start and transcript end.  This constitutes
about 30,000 records.  The function run_pedgene examines the first 100
transcripts and prints the results. The first 100 transcripts
identifies one gene with several markers:

```{r,collapse=TRUE}
run_pedgene()
```

You will see many reports
of "No markers in range", but occasionally you will see a listing of a
gene name, markers, and count of 1/1, 1/2 and 2/2 genotypes, viz.

> CEP104 snp24037 520 646 214  
CEP104 snp24039 940 386 54  
CEP104 snp24041 1377 3 0  
CEP104 snp24048 860 460 60  
CEP104 snp24494 789 501 90  
CEP104 snp24499 891 442 47  
CEP104 snp24506 891 442 47  
CEP104 snp24507 789 501 90

The genotypes for these markers, along with the markers info are fed
to the call back function DOpedgene. DOpedgene converts the genotype
nucleotides to the count of the minor allele and then runs pedgene for
several different marker weights. The results are automatically stored
in a data frame with "observations": prefix of chromosome, gene,
number of markers and base pair range followed by Pedgene data: kernel
and burden, value and p-values, four values for each of three
weightings of the markers. It is printed as well when *verbose* is TRUE, viz.

>    chr   gene nvariants   start     end sKernel_BT pKernel_BT sBurden_BT  
1 chr1 CEP104         8 3728644 3773797   31.96998  0.6297541 -0.6496837  
  pBurden_BT sKernel_MB pKernel_MB sBurden_MB pBurden_MB sKernel_UW pKernel_UW  
1  0.5158965   1184.995  0.5138866  -1.209563  0.2264468   222.8737  0.4111136  
  sBurden_UW pBurden_UW  
1  -1.169488   0.242207

It would be better to run pedgene on all the default transcript entries. So type:

```{r,eval=FALSE}
# we will skip this line for the Rmd document production because it takes too long
applyFnToRanges(DOpedgene, ENV$refRanges, ENV$refIndices, envir = ENV)
```

If you run the above test, you will see that gene DISP1 and KIF26B
have some p-values less than 0.01 and AK5 and STL7 less than 0.03.

Alternatively, you may try searching by Gene.  Here the default
transcript database is TxDb.Hsapiens.UCSC.hg19.knownGene from
Bioconductor.  Of course you can change it.  Type:

> setAnnotations("txdb", "genedb")

where "txdb" is a string that is the name of transcript database that was
fetched from Bioconductor, and similarly "genedb" is the name of a Bioconductor
database that maps a gene id from the transcript to an entrez gene id.

```{r,collapse=TRUE}
applyFnToGenes(DOpedgene, genes_arg = c('DISP1', 'KIF26B', 'AK5', 'ST7L'), envir = ENV)
```

NOTE: You will observe that for several genes there are a few
transcripts with the same start values and the same end values.

## Writing a VCF File using the mega2vcf R package

First create the directory "vcfr"; so we can keep all the generated
data in one place and then load the mega2vcf library.

The Mega2VCF function takes an argument which is the prefix used on all the files
it generates. We have chosen to make that a path that includes a
directory. Mega2VCF is explained in greater depth in the
Implementation Section. The Mega2VCF function assumes the environment
is contained in the global variable ENV. If this is not true supply an
additional argument "envir" set to the environment you wish to use.

Note, that Mega2VCF will create the expected .vcf file as well as a
.fam file (we do linkage), a .phe file (phenotypes) and a few
others. These files are listed below.

Creating the VCF file and related files is accomplished by typing:

```{r}
if (! dir.exists("vcfr")) dir.create("vcfr")
library(mega2vcf)
Mega2VCF("vcfr/vcf.01", ENV$markers[ENV$markers$chromosome==1 ,])
list.files("vcfr")
```

Read the package code to see how these files are created from the data frames. 


## Creating a GenABEL gwaa.data-class object using the mega2genabel R package

GenABEL (http://www.genabel.org/) can process PLINK .tped/.tfam files with the function
*convert.snp.tped* to create a gwaa.data-class object.  Currently, we
use this mechanism to convert our data frames to a gwaa.data-class
object.  The argument to Mega2GenABEL is a file name prefix that will
be used to name the .tped/.fam files that are created as an
intermediate to be read by GenABEL (https://cran.r-project.org/web/packages/GenABEL/index.html)
to produce a GenABEL "raw" file.  Then the "raw" file and a generated phenotype
file are processed by *load.gwaa.data* to yield a gwaa.data-class
object. (See the Implementation Section.)

This task is analogous to the previous exercise. Type:

```{r}
library(mega2genabel)
seqsimgwaa = Mega2GenABEL("SEQ")
list.files(pattern = "SEQ.*")
str(seqsimgwaa)
```
You can use any of the GenABEL provided functions on your data, seqsimgwaa.


Note: There are several data formats that Mega2 can read and transform
into a database that are not directly acceptable to GenABEL. Any
M1ega2 database that can be read into R, can be transformed to GenABEL
following this example with seqsimr.db.

# Regression

To verify that the R packages described above have worked correctly,
we can compare their output files to those created by Mega2 itself, to
verify that they are, as expected, identical. This type of testing is
known as regression testing.

## VCF regression

It is time to use the MEGA2.BATCH.vcf file from the mega2rtutorial.
This will use the C++ Mega2 program to populate the vcf directory with
the same set of files that the R program created in the vcfr
directory.  At the Unix command prompt, type:

```{bash,results="hide"}
mkdir vcf
mega2 MEGA2.BATCH.vcf
```

The abbreviated MEGA2.BATCH.vcf file is below (notice that there are
no INPUT FILES, just a database file):

> Input_Database_Mode=2  
Align_Strand_Input=no  
**Output_Path=vcf**  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
**Analysis_Option=VCF**  
Value_Missing_Quant_On_Output=-9  
Value_Missing_Affect_On_Output=-9  
**DBfile_name=seqsimr.db**  
Chromosome_Single=1  
Traits_Combine=1 2 e  
**file_name_stem=vcf**  
human_genome_build=B37  
VCF_output_file_type=1  
VCF_Allele_Order=Original_Order  
Default_Outfile_Names=yes  

You should compare the two directories: vcf and vcfr. Please add the
-w flag to the diff command. This causes the comparison to ignore
differences in white space. You may may notice that expressions that
contain dates that are different between the two directories.  Type:

```{bash}
diff -wrs vcf vcfr
```

## GenABEL regression

This test is a bit more complicated. First, we load GenABEL and access its data:

```{r}
library(GenABEL)
data(srdta)
```

Then we load the mega2genabel library and dump the GenABEL data as
PLINK .ped/.map/.phe files

```{r}
library(mega2genabel)
dmpPed()
```

Then we switch to the Unix command line and run:

```{bash, results="hide"}
mega2 MEGA2.BATCH.srdta
```

### (abbreviated contents of MEGA2.BATCH.srdata file

> **Input_Database_Mode=1**  
**Input_Format_Type=4**  
**Input_Pedigree_File=srdta.ped**  
**Input_PLINK_Map_File=srdta.map**  
**Input_Phenotype_File=srdta.phe**  
Output_Path=.  
Input_Path=.  
PLINK_Args= --missing-phenotype -9 --trait default  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
Value_Marker_Compression=1  
**Analysis_Option=Dump**  
Value_Missing_Quant_On_Input=-9.000000  
Value_Missing_Affect_On_Input=-9  
Count_Genotypes=4  
Count_Halftyped=no  
Value_Genetic_Distance_Index=0  
Value_Genetic_Distance_SexTypeMap=0  
Value_Base_Pair_Position_Index=1  
Default_Reset_Invalid=no  
**DBfile_name=srdta.db**  
Default_Outfile_Names=yes  

to produce the Mega2 database srdta.db.


Then we go back into R and type:

```{r}
ENV = read.Mega2DB("srdta.db")
mega = Mega2GenABEL("SRD")
list.files(pattern = "SRD.*")
```

*mega* and *srdta* should be the same gwaa.data-class object. You can
compare them however you prefer.  But, Mega2GenABELtst can be used to
compare the phenotype and genotype data, viz. in R type:

```{r}
Mega2GenABELtst()
```

# Next steps

If you are left with questions as to how you can use Mega2R for your own research, you can:

1. Read our paper "The Mega2R suite of R packages: tools for accessing
and processing common genetic data formats in R", which extends this
document by providing implementation details.

2. Read the source for the packages if you understand R; the code is not particularly subtle.

3. Write to the Mega2 discussion group: https://groups.google.com/forum/#!forum/mega2-users
