---
title: "Example usage of Mega2R"
author: "Robert V. Baron and Daniel E. Weeks"
date: "`r format(Sys.time(), '%B %d, %Y, %R')`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
require(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

# Load Libraries

```{r load_libraries,message=FALSE}
```
# Tutorial directory and files 

The files you will need for this tutorial are in the mega2rtutorial directory.  You should **cd** there to do these exercises.:

> You should see the following files:  
........................  | ....................  | ...............  
--------------------------|-----------------------|---------------  
[1] "MEGA2.BATCH.seqsimr" |  "MEGA2.BATCH.srdta"  |  "MEGA2.BATCH.vcf"      
[4] "Mega2r.map"          |  "Mega2r.ped"         |  "mega2r.Rmd"   
[7] "mega2r.pdf"          |  "seqsimr.db"         |    

```{r}
# Print the working directory
getwd()
# Print the files
list.files(pattern = "mega2.*|seqsimr.db", ignore.case = TRUE)
```

# Installation

To run these exercises, you should install the CRAN packages *mega2r*, *mega2vcf*, *mega2genabel* and *mega2pedgene*. Once they are in CRAN, the preferred way to install will be to use the install.packages() command. For now, we prefer to use devtools in R; as in 

```{r,cache=TRUE}
devtools::install("../mega2r")
devtools::install("../mega2pedgene")
devtools::install("../mega2vcf")
devtools::install("../mega2genabel")
```

Certainly, 

> R CMD install mega2r  
R CMD install mega2pedgene  
R CMD install mega2vcf  
R CMD install mega2genabel  

will work, too. It may generate warnings about non-package related items in the tree. All these packages may fetch other dependent CRAN packages. We also require Bioconductor Annotations that must be loaded manually. The first line loads the Bioconductor loader and the next two lines load two annotation databases. One for gene transcript locations and one to map gene names to the entrez gene IDs. (Note: You may choose a different transcript database from Bioconductor or construct one of your own.) Please type in R:

```{r,cache=TRUE}
source("https://bioconductor.org/biocLite.R")
biocLite("TxDb.Hsapiens.UCSC.hg19.knownGene") 
biocLite("org.Hs.eg.db")
```


# Example Data

We have chosen to use the SeqSIMLA2 program to generate a data set for us.  We needed to sub-sample the data down to 1,380 people and 1,000 markers to make the size manageable. These data will be used to illustrate Mega2 and Mega2R operations that follow.

Note: This simulator only produces markers on chromosome 1.

# Using Mega2R packages

## Creating a Mega2 database

We have provided files in the mega2rtutorial directory, that contain the data from the simulation. These files are in PLINK ped format data for 1,380 samples with 1,000 markers:

* seqsimr.ped
* seqsimr.map

You will need to obtain the Mega2 program https://watson.hgen.pitt.edu/register/. First, you will invoke Mega2 on your data. To make matters simple, we will use a pre-constructed Mega2 batch file to automate the processing by Mega2. To run Mega2 to process and create the SQLite3 database 'seqsimr.db', we issue this command at the Unix prompt:

```{bash}
mega2 MEGA2.BATCH.seqsimr
```

If you do not provide the argument, Mega2 will proceed to ask a series of questions to collect the needed information to produce a database. In addition, it will create a Mega2.BATCH file, we've suggested you use. You can look at the "Quick Start" documentation https://watson.hgen.pitt.edu/docs/mega2_html/mega2.html Section 4. to better understand the interactive process.   

This MEGA2.BATCH.seqsimr file begins with a rather long comment indicating the keyword values that may be set and their default value. Toward the end of the file, we set the inputs as **Mega2r.ped** and **Mega2r.map**, indicate the input is PLINK ped format with parameters, and indicate that Mega2 should produce a database called **seqsimr.db**, etc. (These particular items are in bold face text below.)

> **Input_Database_Mode=1**  
**Input_Format_Type=4**  
**Input_Pedigree_File=Mega2r.ped**  
**Input_PLINK_Map_File=Mega2r.map**  
Output_Path=.  
Input_Path=.  
**PLINK_Args= --cM --missing-phenotype -9 --trait default**  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
Value_Marker_Compression=1  
**Analysis_Option=Dump**  
Value_Missing_Quant_On_Input=-9.000000  
Value_Missing_Affect_On_Input=-9  
Count_Genotypes=4  
Count_Halftyped=no  
Value_Genetic_Distance_Index=0  
Value_Genetic_Distance_SexTypeMap=0  
Value_Base_Pair_Position_Index=1  
Default_Reset_Invalid=no  
**DBfile_name=seqsimr.db**  
Default_Outfile_Names=yes  

You will eventually have to run Mega2 to convert your data.  But if you want to temporarily avoid the above processing, notice that in the mega2rtutorial directory, there is a seqsimr.db file.


## Reading a Mega2 database into R

After you have the database, start up the R program. The first R commands you should type are:

```{r,eval=FALSE}
library(mega2r)
ENV = read.Mega2DB("seqsimr.db", verbose = TRUE)
```
The first argument should be the path to the database. Providing the optional argument, verbose, causes the read function to summarize the tables created, their fields and their sizes. Finally, an "R environment" that contains the database tables is returned. (If you are unfamiliar with environments, you can think of them as data frames. If fact, ENV$locus_table will access the locus_table variable from ENV as though fetching an "observation" from a data frame. The difference is when you change a data frame passed to a function, the change does not affect the original data frame. Only the function's local value is changed; ALL changes are forgotten when the function exits. If you change the data in an environment passed to a function, the change is permanent.) 
```{r}
library(mega2r)
ENV = read.Mega2DB("seqsimr.db", verbose = TRUE)
```

We need to make two observations that apply to all Mega2R functions:  The first is that when verbose is set in the initial read.Mega2DB, the value will be remembered.  It may be used by any subsequent function.  If verbose is TRUE, Mega2R functions will possibly print out far more diagnostic information than you might need, especially on subsequent runs.  The second is that all Mega2R functions that do not return an environment, need to have an environment supplied as an argument.  There are two ways to accomplish this.  If you read "seqsimr.db" into the variable *seqsimr*, then you could supply as the named argument **envir** the value *seqsimr* in the Mega2R function as in:

> showMega2ENV(envir = seqsimr)

The second choice is a bit of a "hack" but it is very convenient. Every Mega2R function (that does not return an environment) has a named argument defined as

> envir = ENV

as in

> showMega2ENV = function(envir = ENV) { ... }

The code  above, sets global variable **ENV** to the local envir variable.  And the argument evaluation mechanism of R will supports this useage;  even though **ENV** is not known until runtime.

### Back to more examples.

The ls function will show you all the variables in an environment. (You probably used it without arguments to show you the variables in the .GlobalEnv.) Type:

```{r}
ls(ENV)
```


### A more informative overview of the data base can be seen by:

```{r}
showMega2ENV() 
```

### Use standard R commands to examine the created data frames
```{r}
str(ENV$locus_table)
```

## Running pedgene using the mega2pedgene R package

mega2pedgene is the exception that proves the rule. Rather than use
the read.Mega2DB function described above and as used by all the other
packages, we use init_pedgene. This function calls a utility function
created for read.Mega2DB. Then it creates, edits, and rewrites the
**fam** data frame to purge persons with unknown case/control
status. (**fam** merges data from the pedigree_table, person_table and phenotype_table
as you might expect.)
Finally, it calculates and stores some values that will be
used repeatedly. viz:

```{r}
library(mega2pedgene)
ENV = init_pedgene("seqsimr.db", verbose = T)
```

Mega2R has an internal default list of the chromosome and base pair ranges for a number of gene transcripts.
These transcripts come from the UCSC Genome Browser reference
assembly GRCH37.  The list was further modified to eliminate multiple records of
the same gene with the exact same transcript start and transcript end.
This constitutes about 30,000 records.
The function run_pedgene examines the first 100 transcripts and prints the results. However, our test data is from part of the first chromosome only; the first 100 transcripts find no markers to run pedgene on.
```{r,eval=FALSE}
run_pedgene()
```

When we perform more thorough runs, below, you will see many reports of "No markers in range", but occasionally you will see a listing of a gene name, markers, and count of 1/1, 1/2 and 2/2 genotypes, viz.

> CEP104 snp24027 1280 97 3  
CEP104 snp24788 1375 5 0 

The genotypes for these markers, along with the markers info are fed to the call back function DOpedgene. DOpedgene converts the genotype nucleotides to the count of the minor allele and then runs pedgene for several different marker weights. The results are automatically stored in a file, "k_Schaid_rare.txt". Lines in that file have a prefix of chromosome, gene, number of markers and base pair range followed by 6 p-values. Two for each of three weightings of the markers. It is printed as well.

> chr1 CEP104 2 3728644 3773797 0.8939136 0.8047689 0.8155284 0.6827252 0.9473635 0.9365409 1 
It would be better to run pedgene on all the default transcript entries. So type:

```{r,eval=FALSE}
# we will skip this line for the Rmd document production because it takes too long
applyFnToRanges(DOpedgene, ENV$refRanges, ENV$refIndices, envir = ENV)
```
If you run the above test,
you will see that gene ROR1 has some p-values less than 0.01, FAM129A less than 0.02, and NEK7 and IGSF21 less than 0.03. If we knew this fact to start with, and hence these were the only genes we were interested in.

NOTE: You will observe that for several genes there are a few transcripts with different start and end values.

A shorter test would be

```{r}
# turn off verbose
ENV$verbose = FALSE
applyFnToRanges(DOpedgene, ENV$refRanges[1:300,], ENV$refIndices, envir = ENV)
```

Alternatively, you may try searching by Gene.  Here the default transcript database is TxDb.Hsapiens.UCSC.hg19.knownGene"
from Bioconductor.  Of course you can change it.

```{r}
applyFnToGenes(DOpedgene, genes_arg = c('ROR1', 'FAM129A', 'NEK7', 'IGSF21'), envir = ENV)
```

or

```{r}
applyFnToGenes(DOpedgene, genes_arg = c('GDAP2'), envir = ENV)
```

## Writing a VCF File using the mega2vcf R package

First create the directory "vcfr"; so we can keep all the generated data in one place and then load the library.

Mega2VCF takes an argument which is the prefix used on all the files it generates. We have chosen to make that a path that includes a directory. Mega2VCF is explained in greater depth in the Implementation Section. The Mega2VCF function assumes the environment is contained in the global variable ENV. If this is not true supply an additional argument "envir" set to the environment you wish to use.

Note, that Mega2VCF will create the expected .vcf file as well as a .fam file (we do linkage), a .phe file (phenotypes) and a few others. These files are listed below.

Creating the VCF file and related files is accomplished by typing:

```{r}
if (! dir.exists("vcfr")) dir.create("vcfr")
library(mega2vcf)
Mega2VCF("vcfr/vcf.01", ENV$markers[ENV$markers$chromosome==1 ,])
list.files("vcfr")
```
Read the package code to see how these files are created from the data frames. 


## Creating a GenABEL gwaa.data-class object using the mega2genabel R package

GenABEL can process PLINK .tped/.tfam files with the finction *convert.snp.tped* to create a gwaa.data-class object.  Currently, we use this
mechanism to convert our data frames to a gwaa.data-class object.
The argument to Mega2GenABEL is a file name prefix that will be used to name the .tped/.fam files that are created as an intermediate to be read by GenABEL http://www.genabel.org/ https://cran.r-project.org/web/packages/GenABEL/index.html to produce a GenABEL "raw" file. 
Then the "raw" file and a generated phenotype file are processed by *load.gwaa.data* to yield a gwaa.data-class object. (See the Implementation Section.) 

This task is analogous to the previous exercise. Type:

```{r}
library(mega2genabel)
seqsimgwaa = Mega2GenABEL("TMP")
list.files(pattern = "TMP.*")
str(seqsimgwaa)
```
You can use any of the GenABEL provided functions on your data, seqsimgwaa.


Note: There are several data formats that Mega2 can read and transform into a database that are not directly acceptable to GenABEL. Any M1ega2 database that can be read into R, can be transformed to GenABEL following this example with seqsimr.db.

# Regression

To verify that the R packages described above have worked correctly, we can compare their output files to those created by Mega2 itself, to verify that they are, as expected, identical. This type of testing is known as regression testing.

## VCF regression

It is time to use the MEGA2.BATCH.vcf file from the mega2rtutorial.
This will use the C++ Mega2 program to populate the vcf directory with the same set of files that the R program created in the vcfr directory.
At the command prompt, type:

```{bash,eval=FALSE}
mkdir vcf
mega2 MEGA2.BATCH.vcf
```

The abbreviated MEGA2.BATCH.vcf file is below (notice that there are no INPUT FILES, just a database file):

> Input_Database_Mode=2  
Align_Strand_Input=no  
**Output_Path=vcf**  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
**Analysis_Option=VCF**  
Value_Missing_Quant_On_Output=-9  
Value_Missing_Affect_On_Output=-9  
**DBfile_name=seqsimr.db**  
Chromosome_Single=1  
Traits_Combine=1 2 e  
**file_name_stem=vcf**  
human_genome_build=B37  
VCF_output_file_type=1  
VCF_Allele_Order=Original_Order  
Default_Outfile_Names=yes  

```{bash}
mkdir vcf
mega2 MEGA2.BATCH.vcf
```

You should compare the two directories: vcf and vcfr. Please add the -w flag to the diff command. This causes the comparison to ignore differences in white space. You may may notice that expressions that contain dates that are different between the two directories.  Type:

```{bash}
diff -wrs vcf vcfr
```

## GenABEL regression

This test is a bit more complicated. First, we load GenABEL and access its data:

```{r}
library(GenABEL)
data(srdta)
```

We also load the mega2genabel library and dump the GenABEL data as PLINK .ped/.map/.phe files

```{r}
library(mega2genabel)
dmpPed()
```

Then we switch to the command line and run:

```{bash,eval=FALSE}
mega2 MEGA2.BATCH.srdta
```

### (abbreviated contents of MEGA2.BATCH.srdata file

> **Input_Database_Mode=1**  
**Input_Format_Type=4**  
**Input_Pedigree_File=srdta.ped**  
**Input_PLINK_Map_File=srdta.map**  
**Input_Phenotype_File=srdta.phe**  
Output_Path=.  
Input_Path=.  
PLINK_Args= --missing-phenotype -9 --trait default  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
Value_Marker_Compression=1  
**Analysis_Option=Dump**  
Value_Missing_Quant_On_Input=-9.000000  
Value_Missing_Affect_On_Input=-9  
Count_Genotypes=4  
Count_Halftyped=no  
Value_Genetic_Distance_Index=0  
Value_Genetic_Distance_SexTypeMap=0  
Value_Base_Pair_Position_Index=1  
Default_Reset_Invalid=no  
**DBfile_name=srdta.db**  
Default_Outfile_Names=yes  

to produce the Mega2 database srdta.db.

```{bash}
mega2 MEGA2.BATCH.srdta
```

Then we go back into R and type:

```{r}
ENV = read.Mega2DB("srdta.db")
mega = Mega2GenABEL("TMP2")
```

*mega* and *srdta* should be the same gwaa.data-class object. You can compare them however you prefer.
But, Mega2GenABELtst can be used to compare the phenotype and genotype data, viz. in R type:

```{r}
Mega2GenABELtst()
```

# Next steps

If you are left with questions as to how you can use Mega2R for your own research, you can:

1. Read our paper "The Mega2R suite of R packages: tools for accessing and processing common genetic data formats in R", which extends this document by providing implementation details.

2. Read the source for the packages if you understand R; the code is not particularly subtle.

3. Write to the Mega2 discussion group: https://groups.google.com/forum/#!forum/mega2-users
