---
title: "Mega2pedgene Vignette"
author: "Robert V. Baron and Daniel E. Weeks"
date: "`r format(Sys.time(), '%B %d, %Y, %R')`"
output:
  rmarkdown::html_vignette:
  pdf_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Mega2pedgene Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```
# Loading genetic annotation data
Bioconductor Annotations must be loaded
manually. The first line (below) loads the Bioconductor loader and the next
two lines load two annotation databases. One for gene transcript
locations and one to map gene names to the entrez gene IDs. (Note: You
may choose a different transcript database from Bioconductor or
construct one of your own.)

Please type in R:

```
source("https://bioconductor.org/biocLite.R")
biocLite("TxDb.Hsapiens.UCSC.hg19.knownGene") 
biocLite("org.Hs.eg.db")
```
You type the above code only when you install/upgrade the R executable.

The Mega2r package presumes that the two names shown above are used for the Annotation Databases.
If you desire to use a different database, you need to load it with biocLite and indicate the name by running the 
command:
```
setAnnotations("transcript database name", "gene map database name")
```
where "transcript database name" is a string that is the name of transcript database that was
fetched from Bioconductor, and similarly "gene map database name" is the name of a Bioconductor
database that maps a gene id from the transcript to an entrez gene id.
This command must be issued whenever R is started and the "mega2r" library has been loaded.

# Running pedgene using the mega2pedgene R package

## Loading a Mega2 database

Rather than use
the read.Mega2DB function described elsewhere and used by all the other
packages, This package uses *init_pedgene* to read the Mega2 database.
This function calls a utility function
shared with read.Mega2DB. Then it creates, edits, and rewrites the family data -- 
data frame, **fam** .  (**fam** [merges data from the pedigree_table, person_table and
phenotype_table as you might expect.)
*init_pedgene* purges persons with unknown case/control
status which is necessary for the pedgene calculation.   Finally, *init_pedgene* calculates
some values that will be used repeatedly and stores them in the environment, ENV.

```{r mega2pedgene}
library(mega2pedgene)
# ENV = init_pedgene("seqsimr.db", verbose = T)
# or to use the embeded tutorial data
library(mega2rtutorial)
database_file = system.file("exdata", "seqsimr.db", package="mega2rtutorial")
ENV = init_pedgene(database_file, verbose = T)
```

## Running pedgene on transcripts

Mega2R has an internal default list of the chromosome and base pair
ranges for a number of gene transcripts.  These transcripts come from
the UCSC Genome Browser reference assembly GRCH37.  The list was
further modified to eliminate multiple records of the same gene with
the exact same transcript start and transcript end.  This constitutes
about 29,000 records.  You may load you own range set instead of the default.
You create a data frame that contains at least a name "observation", a chromosome
"observation", a start position "observation" and an end position "observation".
You create an integer vector that contains the column numbers of the chromosome
"observation", the start position "observation" and the end position "observation".
These two become the arguments to the *setRanges* function, viz.
```
        setRanges(Transcripts, Offsets)
```

By default, the function run_pedgene examines the first 100
transcripts and prints the results. For this database, the first 100 transcripts
identifies one useable transcript with several markers:  (To make this exercise run 
faster, we noticed that the data appeared at transcript 54; so we will
restrict pedgene to a small range of transcripts around 54, viz. 50 through
60.)

```{r run_pedgene,collapse=TRUE}
run_pedgene(gs = 50:60)
```

You will see many reports
of "No markers in range", because the database is a subrange of chromosome 1 whereas the transcripts span
the entire genome.
Occasionally you will see a listing of a
gene name, markers, and count of 1/1, 1/2 and 2/2 genotypes, viz.

```
CEP104 snp24037 520 646 214  
CEP104 snp24039 940 386 54  
CEP104 snp24041 1377 3 0  
CEP104 snp24048 860 460 60  
CEP104 snp24494 789 501 90  
CEP104 snp24499 891 442 47  
CEP104 snp24506 891 442 47  
CEP104 snp24507 789 501 90
```

The genotypes for these markers, along with the markers info are passed
to the call back function DOpedgene. DOpedgene converts the genotype
nucleotides to the count of the minor allele and then runs pedgene for
several different marker weights. The results are automatically stored
in a data frame with "observations": prefix of chromosome, gene,
number of markers and base pair range followed by Pedgene data: kernel
and burden, value and p-values, four values for each of three
weightings of the markers.
These data are saved in the data frame, **pedgene_results**, in the environment, ENV.
They are also printed when *verbose* is TRUE, viz.

```
   chr   gene nvariants   start     end sKernel_BT pKernel_BT sBurden_BT  
1 chr1 CEP104         8 3728644 3773797   31.96998  0.6297541 -0.6496837  
  pBurden_BT sKernel_MB pKernel_MB sBurden_MB pBurden_MB sKernel_UW pKernel_UW  
1  0.5158965   1184.995  0.5138866  -1.209563  0.2264468   222.8737  0.4111136  
  sBurden_UW pBurden_UW  
1  -1.169488   0.242207
```

We should run pedgene on all the transcript entries. So type:

```{r,eval=FALSE}
# we will skip this line for the Rmd document production because it takes too long
applyFnToRanges(DOpedgene, ENV$refRanges, ENV$refIndices, envir = ENV)
```

If you run the above test, you will see that gene DISP1 and KIF26B
have some p-values less than 0.01 and AK5 and STL7 less than 0.03.

## Running pedgene on selected genes

You may try searching for transcripts of specific genes.  Here the default
transcript database is TxDb.Hsapiens.UCSC.hg19.knownGene from
Bioconductor.  Of course you can change it.  Type:

```
setAnnotations("txdb", "genedb")
```

where "txdb" is a string that is the name of transcript database that was
fetched from Bioconductor, and similarly "genedb" is the name of a Bioconductor
database that maps a gene id from the transcript to an entrez gene id.

We leave the command below as an exercise.  It runs a bit slowly.  It needs to find
all  the transcripts for each gene; then it needs to find all the markers between each pair
of transcript start/end ranges; and finally it needs to do a small computation on the genotypes
from each set of markers.
```{r,eval=FALSE,collapse=TRUE}
applyFnToGenes(DOpedgene, genes_arg = c('DISP1', 'KIF26B', 'AK5', 'ST7L'), envir = ENV)
```

But let us do at least a few genes:

```{r,eval=TRUE,collapse=TRUE}
applyFnToGenes(DOpedgene, genes_arg = c('DISP1', 'AK5'), envir = ENV)
```

NOTE: You will observe that for several genes there are a few
transcripts with the same or overlapping start values and end values.
