---
title: "Mega2VCF Vignette"
author: "Robert V. Baron and Daniel E. Weeks"
date: "`r format(Sys.time(), '%B %d, %Y, %R')`"
output:
  rmarkdown::html_vignette:
  pdf_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Mega2VCF Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

# Writing a VCF File using the mega2vcf R package

First load the mega2vcf library and then create the directory "vcfr"; so we can keep all the generated
data in one place.

```{r mega2vcflib}
library(mega2vcf)

if (! dir.exists("vcfr")) dir.create("vcfr")
```

The Mega2VCF function takes an argument which is the prefix used on all the files
generated. We will make the prefix path include the
directory, *vcfr*.  The Mega2VCF function assumes the Mega2 data frames
are contained in the global variable, **ENV**. If this is not the case, supply a
named argument, *envir*, to indicate the environment you wish to use.

Note, that Mega2VCF will create a .vcf file as well as a
.fam file (we do linkage), a .phe file (phenotypes) and a few
others. These files are listed below.

To create the VCF file and related files, type:

```{r mega2vcf}
# ENV = read.Mega2DB("seqsimr.db")

# or to use the embeded tutorial data
library(mega2rtutorial)
database_file = system.file("exdata", "seqsimr.db", package = "mega2rtutorial")
ENV = read.Mega2DB(database_file)

Mega2VCF("vcfr/vcf.01", ENV$markers[ENV$markers$chromosome==1 ,])
```

Note: This command only creates files for chromosome 1.  You can filter ENV$markers as needed.  
Note: If the second argument is not present, all the markers are processed.

The generated files are listed below:
```{r}
list.files("vcfr")
```

You can read the R code to see how these files are created from the data frames. 


# VCF regression

If you have a executable copy of Mega2, you can run the code below using
the MEGA2.BATCH.vcf file and using the database from the *mega2rtutorial* package.

To copy the batch file and database from the tutorial to the current directory type:
```{r, eval=FALSE}
library(mega2rtutorial)
dump_mega2rtutorial_data()
```

The C++ Mega2 program can populate the vcf directory with
the same set of files that the R program, *Mega2VCF*, created in the vcfr
directory.  So, at the Unix command prompt, type:

```
mkdir vcf
mega2 MEGA2.BATCH.vcf
```

**Note: To make this tutorial only dependent on R, the above code is not actually run.  And the
results, shown below, were captured from an environment where we had both R and the Mega2
executable available.**

The output should be similar to that below (except for timestamps):

```
## ==========================================================
##                           MEGA2 4.9.2
##      Copyright (C) 1999-2017 Robert Baron, Justin R. Stickel, Charles P. Kollar,
##      Nandita Mukhopadhyay, Lee Almasy, Mark Schroeder, William P. Mulvihill,
##      Daniel E. Weeks, and University of Pittsburgh
## 
##      Last updated: Jun 13 2017, 09:36:42 , valid until June 15, 2018.
##      Compiled with gcc version 4.2.1 Compatible Apple LLVM 8.0.0 (clang-800.0.42.1)
## 
##      Mega2 comes with ABSOLUTELY NO WARRANTY.
##      See LICENSE.txt for terms of copying, modifying & redistributing Mega2.
## ==========================================================
## NOTE: For humans, chromosome 23 codes for X, 24 codes for Y and 25 codes for XY.
## 
## Run date:                  2017-7-24-10-05
## 
## Running Mega2 in batch mode from MEGA2.BATCH.vcf.
## Analysis option read in from batch file.
## Chromosome(s) and markers read in from batch file.
## Trait selection(s) read in from batch file.
## ==========================================================
## Analysis Class: VCF.
## Quantitative Output Missing Value     "-9"
## Affection    Output Missing Value     "-9"
## Reading SQLite3 DB from file "seqsimr.db"
## ===========================================================
## The path to this SQLite3 database is seqsimr.db.
## This database was created using Mega2 version 4.9.2.
## This database was created using   SQLite3 3.9.2 on 2017-7-24-10-03.
## This database was processed using SQLite3 3.9.2 on 2017-7-24-10-05.
## This database was created from PLINK PED format data using the following files:
## 	       Pedigree file   Mega2r.ped
## 	      PLINK Map file   Mega2r.map
## This database contains:
## 	1380 persons (20 pedigrees)
## 	1000 markers
## 	1 trait
## 	genetic distance(map name/type) "Map"/kosambi, Sex map type AVERAGED_MAP
## 	base pair distance(map name) "BP"
## 
## ==========================================================
## 1 trait locus 
##       1 Affection status locus: 
##                 default
## ===========================================================
## Selected map Map.
## Selected chromosome 1
## Output will combine markers and the following selected traits:
##                 default [MARKERS]
## After selecting traits and covariates
## 1 trait locus 
##       1 Affection status locus: 
##                 default
## ===========================================================
## Pedigree statistics after selecting chromosomes and marker loci:
## Input pedigree file is in post-makeped format.
##                                                   Marker Genotypes
##                                                   Fully    Half
##      Pedigrees   People   Males   Females         Typed    Typed     Total
## TOTAL       20     1380     620       760       1380000        0    1380000
## Typed       20     1380     620       760
## Untyped      0        0       0         0
## ===========================================================
## Mega2 created the following file(s) for VCF Format:
## VCF file created using allele ordering setting: Original_Order
##         VCF format file:        vcf/vcf.01.vcf
##         VCF pedigree file:      vcf/vcf.01.fam
##         VCF phenotype file:     vcf/vcf.01.phe
##         VCF map file:           vcf/vcf.01.map
##         VCF freq file:          vcf/vcf.01.freq
##         VCF pen file:           vcf/vcf.01.pen
## 
## SQLite3 database "seqsimr.db" was processed to generate this output.
## Output is in vcf
## ===========================================================
## See run summaries in directory 2017-7-24-10-05 
##    MEGA2.LOG, MEGA2.ERR, MEGA2.KEYS
## The script 'mega2log2html.pl' exited normally.
## To view the HTML-formatted run summaries, open
## /Users/rbaron/mega2/bb/srcdir/R/mega2rtutorial/vignettes/2017-7-24-10-05/MEGA2run.html
## in a web browser.
## ===========================================================
## If you use Mega2 as part of a published work, please reference 
##  Baron RV, Kollar C, Mukhopadhyay N, Weeks DE
##  Mega2: validated data-reformatting for linkage and association analyses
##  Source Code for Biology and Medicine.2014, 9:26
##  DOI: 10.1186/s13029-014-0026-y
## as well as the version used, which is currently Version 4.9.2
## ===========================================================
```


The abbreviated MEGA2.BATCH.vcf file is below. (The initial comment section
is not shown and important lines are
shown in a bold typeface.)
Notice that there are no input files, other than a database file. 

> Input_Database_Mode=2  
Align_Strand_Input=no  
**Output_Path=vcf**  
Input_Untyped_Ped_Option=2  
Input_Do_Error_Sim=no  
AlleleFreq_SquaredDev=999999999.000000  
**Analysis_Option=VCF**  
Value_Missing_Quant_On_Output=-9  
Value_Missing_Affect_On_Output=-9  
**DBfile_name=seqsimr.db**  
Chromosome_Single=1  
Traits_Combine=1 2 e  
**file_name_stem=vcf**  
human_genome_build=B37  
VCF_output_file_type=1  
VCF_Allele_Order=Original_Order  
Default_Outfile_Names=yes  

You should compare the two directories: vcf and vcfr. Please add the
-w flag to the **diff** command. This causes the comparison to ignore
differences in white space. Lines that
contain dates may be different between the two directories.  At a Unix
command prompt, type:

```
diff -wrs vcf vcfr
```

**Note: To make this tutorial only dependent on R, the above code is not actually run.  And the
results, shown below, were captured from an environment where we had both R and the Mega2
executable available.**

It should show:

```
## Files vcf/vcf.01.fam and vcfr/vcf.01.fam are identical
## Files vcf/vcf.01.freq and vcfr/vcf.01.freq are identical
## Files vcf/vcf.01.map and vcfr/vcf.01.map are identical
## Files vcf/vcf.01.pen and vcfr/vcf.01.pen are identical
## Files vcf/vcf.01.phe and vcfr/vcf.01.phe are identical
## Files vcf/vcf.01.vcf and vcfr/vcf.01.vcf are identical
```
